---
title: "Rolling-Origin Backtest — Kinsa Models"
output: html_document
date: "2026-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, message=FALSE, warning=FALSE}
library(rjags)
library(tidyverse)
library(coda)
library(tidycensus)
library(HDInterval)
library(scoringRules)
library(patchwork)

# Fit functions
source("R/fit_base_kinsa.R")
source("R/fit_m1_kinsa.R")
source("R/fit_m2_kinsa.R")
source("R/fit_m3_kinsa.R")
```

## Load data once

Same data-loading pipeline as `Base_Model_Kinsa.Rmd`. Done once up-front so
the rolling-origin loop just slices this dataframe instead of re-downloading.

```{r load_data, message=FALSE, warning=FALSE}
# -------- FILL IN: local path to your Kinsa CSV --------
kinsa_path <- "/Users/tobysalmon/Desktop/kinsa_michigan_weekly.csv"   
# -------------------------------------------------------

kinsa_state <- "Michigan"
fips        <- "26"

data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz")
kinsa_raw <- read_csv(kinsa_path)

data <- data %>%
  filter(geography == fips, age == "Total")

kinsa_data <- kinsa_raw %>%
  filter(region_name == kinsa_state) %>%
  mutate(time = as.Date(time)) %>%
  select(time, percent_ill) %>%
  arrange(time)

state_pre <- get_estimates(
  geography = "state", state = fips,
  product = "population", vintage = 2019, time_series = TRUE
) |>
  filter(variable == "POP") |>
  mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |>
  arrange(year) |> pull(value)

state_pop <- get_estimates(
  geography = "state", state = fips,
  product = "population", vintage = 2025, time_series = TRUE
) |>
  filter(variable == "POPESTIMATE", year %in% c(2020:2025)) %>%
  arrange(year) %>% pull(value)

state_pop_estimates <- tibble(
  year = 2018:2026,
  population = c(
    state_pre[1], state_pre[2],
    state_pop[1], state_pop[2], state_pop[3],
    state_pop[4], state_pop[5], state_pop[6], state_pop[6]
  )
)

data_with_pop <- data %>%
  mutate(year = year(time)) %>%
  left_join(state_pop_estimates, by = "year") %>%
  mutate(
    covid_count = round(rate_covid * population / 100000),
    flu_count   = round(rate_flu   * population / 100000),
    rsv_count   = round(rate_rsv   * population / 100000)
  ) %>%
  inner_join(kinsa_data, by = "time") %>%
  arrange(time)

cat("Merged panel:", nrow(data_with_pop), "weeks\n")
cat("Date range:", as.character(min(data_with_pop$time)), "to",
    as.character(max(data_with_pop$time)), "\n")
```

## Backtest configuration

```{r backtest-config}
# Number of rolling-origin as-of dates to evaluate.
# Start small (5-10) while debugging, then bump to 20+ for the paper.
n_as_of <- 10

max_time <- max(data_with_pop$time)
as_of_dates <- seq(max_time - weeks(n_as_of - 1), max_time, by = "week")

# Registry of all models to backtest
model_registry <- list(
  base_kinsa = fit_base_kinsa,
  m1_kinsa   = fit_m1_kinsa
  #m2_kinsa   = fit_m2_kinsa,
  #m3_kinsa   = fit_m3_kinsa
)

results_path <- "results/backtest_predictions.csv"
dir.create("results", showWarnings = FALSE, recursive = TRUE)

cat("Will fit", length(model_registry), "model(s) x", length(as_of_dates),
    "as-of dates =", length(model_registry) * length(as_of_dates), "total fits\n")
cat("As-of dates:\n")
print(as_of_dates)
```

## Run the backtest (resumable / cached)

Each (model, as_of) fit is written to `results/backtest_predictions.csv`
immediately after it finishes. If you re-run this chunk, already-completed
fits are skipped — so a crash or interruption just means picking up where
you left off.

```{r run-backtest, message=TRUE, warning=FALSE}
log_msg <- function(...) {
  cat(..., "\n", sep = "")
  flush.console()
}

existing <- if (file.exists(results_path)) {
  read_csv(results_path, show_col_types = FALSE) %>%
    mutate(as_of = as.Date(as_of), target_date = as.Date(target_date))
} else {
  tibble()
}

for (mname in names(model_registry)) {
  fit_fn <- model_registry[[mname]]

  for (i in seq_along(as_of_dates)) {
    d <- as.Date(as_of_dates[i])

    already_done <- nrow(existing) > 0 &&
      any(existing$model == mname & existing$as_of == d)

    if (already_done) {
      log_msg(sprintf("[%s | %s] cached, skipping", mname, format(d)))
      next
    }

    t0 <- Sys.time()
    log_msg(sprintf("[%s | %s] fitting...", mname, format(d)))

    preds <- tryCatch(
      fit_fn(data_with_pop, as_of_date = d),
      error = function(e) {
        log_msg("  ERROR: ", conditionMessage(e))
        NULL
      }
    )

    if (is.null(preds)) next

    existing <- bind_rows(existing, preds)
    write_csv(existing, results_path)

    dt <- round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1)
    log_msg(sprintf("  done in %ss (%d rows)", dt, nrow(preds)))
  }
}

results <- read_csv(results_path, show_col_types = FALSE) %>%
  mutate(as_of = as.Date(as_of), target_date = as.Date(target_date))

cat("\nTotal rows in results:", nrow(results), "\n")
```

## Summary metrics

```{r metrics}
metrics <- results %>%
  group_by(model, disease) %>%
  summarise(
    n          = n(),
    MAE        = mean(abs(truth - median)),
    RMSE       = sqrt(mean((truth - median)^2)),
    CRPS       = mean(crps),
    cover_50   = mean(truth >= q25  & truth <= q75),
    cover_95   = mean(truth >= q025 & truth <= q975),
    .groups = "drop"
  ) %>%
  arrange(disease, model)

cat("=== Rolling-origin backtest metrics ===\n")
print(metrics)
```

## Per-week score distributions

Box plots of per-week CRPS across the backtest window — gives a sense of
variability, not just the mean.

```{r crps-boxplot, fig.width=9, fig.height=5}
ggplot(results, aes(x = model, y = crps, fill = model)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.4) +
  facet_wrap(~ disease, scales = "free_y") +
  labs(title = "Per-week CRPS across rolling-origin backtest",
       x = NULL, y = "CRPS (lower is better)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Prediction traces over the backtest window

For each disease, plot the truth (points) against each model's median
prediction with a 50% interval.

```{r prediction-traces, fig.width=10, fig.height=8}
ggplot(results, aes(x = target_date)) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = model), alpha = 0.25) +
  geom_line(aes(y = median, color = model), linewidth = 0.8) +
  geom_point(aes(y = truth), color = "black", size = 2) +
  facet_wrap(~ disease, ncol = 1, scales = "free_y") +
  labs(title = "Rolling-origin nowcasts vs. truth",
       subtitle = "Points = observed, lines = posterior median, ribbon = 50% CI",
       x = NULL, y = "Count") +
  theme_minimal()
```

## Error over time

```{r error-over-time, fig.width=10, fig.height=5}
results %>%
  mutate(abs_error = abs(truth - median)) %>%
  ggplot(aes(x = target_date, y = abs_error, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ disease, scales = "free_y") +
  labs(title = "Absolute error by week", x = NULL, y = "|truth - median|") +
  theme_minimal()
```
