---
title: "Rolling-Origin Backtest — Kinsa Models"
output: html_document
date: "2026-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, message=FALSE, warning=FALSE}
library(rjags)
library(tidyverse)
library(coda)
library(tidycensus)
library(HDInterval)
library(scoringRules)
library(patchwork)

# Fit functions — Kinsa models
source("R/fit_base_kinsa.R")
source("R/fit_m1_kinsa.R")
source("R/fit_m2_kinsa.R")
source("R/fit_m3_kinsa.R")

# Fit functions — ILI models
source("R/fit_base_ili.R")
source("R/fit_m1_ili.R")
source("R/fit_m2_ili.R")
source("R/fit_m3_ili.R")
```

## Load data once

Same data-loading pipeline as `Base_Model_Kinsa.Rmd`. Done once up-front so
the rolling-origin loop just slices this dataframe instead of re-downloading.

```{r load_data, message=FALSE, warning=FALSE}
# -------- FILL IN: local path to your Kinsa CSV --------
kinsa_path <- "/Users/tobysalmon/Desktop/kinsa_michigan_weekly.csv"   
# -------------------------------------------------------

kinsa_state <- "Michigan"
fips        <- "26"

data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz")
ili_data_raw <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/delphi_ili_fluview/standard/data.csv.gz")
kinsa_raw <- read_csv(kinsa_path)

data <- data %>%
  filter(geography == fips, age == "Total")

# ILI data (Delphi FluView)
ili_data <- ili_data_raw %>%
  filter(geography == fips) %>%
  select(time, ili_count = delphi_fluview_num_ili,
         ili_number_patients_tested = delphi_fluview_num_patients) %>%
  arrange(time)

kinsa_data <- kinsa_raw %>%
  filter(region_name == kinsa_state) %>%
  mutate(time = as.Date(time)) %>%
  select(time, percent_ill) %>%
  arrange(time)

state_pre <- get_estimates(
  geography = "state", state = fips,
  product = "population", vintage = 2019, time_series = TRUE
) |>
  filter(variable == "POP") |>
  mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |>
  arrange(year) |> pull(value)

state_pop <- get_estimates(
  geography = "state", state = fips,
  product = "population", vintage = 2025, time_series = TRUE
) |>
  filter(variable == "POPESTIMATE", year %in% c(2020:2025)) %>%
  arrange(year) %>% pull(value)

state_pop_estimates <- tibble(
  year = 2018:2026,
  population = c(
    state_pre[1], state_pre[2],
    state_pop[1], state_pop[2], state_pop[3],
    state_pop[4], state_pop[5], state_pop[6], state_pop[6]
  )
)

data_with_pop <- data %>%
  mutate(year = year(time)) %>%
  left_join(state_pop_estimates, by = "year") %>%
  mutate(
    covid_count = round(rate_covid * population / 100000),
    flu_count   = round(rate_flu   * population / 100000),
    rsv_count   = round(rate_rsv   * population / 100000)
  ) %>%
  inner_join(kinsa_data, by = "time") %>%
  left_join(ili_data, by = "time") %>%
  mutate(ili_count = replace_na(ili_count, 0)) %>%
  arrange(time)

cat("Merged panel:", nrow(data_with_pop), "weeks\n")
cat("Date range:", as.character(min(data_with_pop$time)), "to",
    as.character(max(data_with_pop$time)), "\n")
cat("Kinsa percent_ill NAs:", sum(is.na(data_with_pop$percent_ill)), "\n")
cat("ILI count NAs:", sum(is.na(data_with_pop$ili_count)), "\n")
```

## Backtest configuration

```{r backtest-config}
library(parallel)

# Number of rolling-origin as-of dates to evaluate.
n_as_of <- 100

# Stop the backtest window 3 weeks before the end of the data so that the
# "truth" used for evaluation has had time to mature (see data maturation
# analysis — respnet rates don't converge to final values until ~3 weeks
# after the event week).
max_time <- max(data_with_pop$time) - weeks(3)
as_of_dates <- seq(max_time - weeks(n_as_of - 1), max_time, by = "week")

# Registry of all models to backtest
model_registry <- list(
  base_kinsa = fit_base_kinsa,
  m1_kinsa   = fit_m1_kinsa,
  m2_kinsa   = fit_m2_kinsa,
  m3_kinsa   = fit_m3_kinsa,
  base_ili   = fit_base_ili,
  m1_ili     = fit_m1_ili,
  m2_ili     = fit_m2_ili,
  m3_ili     = fit_m3_ili
)

# Parallelization: 8 of 10 cores (leaves 2 for OS).
n_cores <- 8

results_path <- "results/backtest_predictions.csv"
dir.create("results", showWarnings = FALSE, recursive = TRUE)

cat("Will fit", length(model_registry), "model(s) x", length(as_of_dates),
    "as-of dates =", length(model_registry) * length(as_of_dates), "total fits\n")
cat("Using", n_cores, "cores\n")
cat("Holdout: 3 weeks of RESP-NET, 0 weeks of Kinsa\n")
cat("Backtest window:", as.character(min(as_of_dates)), "to",
    as.character(max_time), "\n")
```

## Run the backtest (parallel, resumable)

Fits run in parallel batches using `parallel::mclapply`. Results are
written to `results/backtest_predictions.csv` after each batch, so a
crash only loses the current batch (not all prior work).

```{r run-backtest, message=TRUE, warning=FALSE}
log_msg <- function(...) {
  cat(..., "\n", sep = "")
  flush.console()
}

existing <- if (file.exists(results_path)) {
  read_csv(results_path, show_col_types = FALSE) %>%
    mutate(as_of = as.Date(as_of), target_date = as.Date(target_date))
} else {
  tibble()
}

# Build grid of all (model, as_of) combinations
grid <- expand.grid(
  model_name = names(model_registry),
  as_of = as_of_dates,
  stringsAsFactors = FALSE
) %>%
  mutate(as_of = as.Date(as_of))

# Remove already-cached fits
if (nrow(existing) > 0) {
  cached <- existing %>% distinct(model, as_of)
  grid <- grid %>%
    anti_join(cached, by = c("model_name" = "model", "as_of" = "as_of"))
}

log_msg(sprintf("%d fits remaining (%d cached)", nrow(grid),
                length(model_registry) * length(as_of_dates) - nrow(grid)))

if (nrow(grid) > 0) {
  # Process in batches so we can save progress periodically
  batch_size <- n_cores * 3   # save after every ~3 rounds of parallel fits
  n_batches <- ceiling(nrow(grid) / batch_size)

  for (b in seq_len(n_batches)) {
    batch_start <- (b - 1) * batch_size + 1
    batch_end   <- min(b * batch_size, nrow(grid))
    batch       <- grid[batch_start:batch_end, ]

    t0 <- Sys.time()
    log_msg(sprintf("Batch %d/%d: fits %d-%d (%d fits on %d cores)",
                    b, n_batches, batch_start, batch_end,
                    nrow(batch), n_cores))

    batch_results <- mclapply(seq_len(nrow(batch)), function(i) {
      mname <- batch$model_name[i]
      d     <- batch$as_of[i]
      fit_fn <- model_registry[[mname]]
      tryCatch(
        fit_fn(data_with_pop, as_of_date = d),
        error = function(e) {
          warning(sprintf("  ERROR [%s | %s]: %s", mname, d, conditionMessage(e)))
          NULL
        }
      )
    }, mc.cores = n_cores)

    # Collect non-NULL results from this batch
    new_preds <- bind_rows(Filter(Negate(is.null), batch_results))

    if (nrow(new_preds) > 0) {
      existing <- bind_rows(existing, new_preds)
      write_csv(existing, results_path)
    }

    dt <- round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1)
    log_msg(sprintf("  batch done in %s min | %d total rows saved",
                    dt, nrow(existing)))
  }
}

results <- read_csv(results_path, show_col_types = FALSE) %>%
  mutate(as_of = as.Date(as_of), target_date = as.Date(target_date))

cat("\nTotal rows in results:", nrow(results), "\n")
cat("Unique (model, as_of) combos:", nrow(distinct(results, model, as_of)), "\n")
```

## Summary metrics

```{r metrics}
metrics <- results %>%
  group_by(model, disease) %>%
  summarise(
    n          = n(),
    MAE        = mean(abs(truth - median)),
    RMSE       = sqrt(mean((truth - median)^2)),
    CRPS       = mean(crps),
    cover_50   = mean(truth >= q25  & truth <= q75),
    cover_95   = mean(truth >= q025 & truth <= q975),
    .groups = "drop"
  ) %>%
  arrange(disease, model)

cat("=== Rolling-origin backtest metrics ===\n")
print(metrics)
```

## Per-week score distributions

Box plots of per-week CRPS across the backtest window — gives a sense of
variability, not just the mean.

```{r crps-boxplot, fig.width=9, fig.height=5}
ggplot(results, aes(x = model, y = crps, fill = model)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.4) +
  facet_wrap(~ disease, scales = "free_y") +
  labs(title = "Per-week CRPS across rolling-origin backtest",
       x = NULL, y = "CRPS (lower is better)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Prediction traces over the backtest window

For each disease, plot the truth (points) against each model's median
prediction with a 50% interval.

```{r prediction-traces, fig.width=10, fig.height=8}
ggplot(results, aes(x = target_date)) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = model), alpha = 0.25) +
  geom_line(aes(y = median, color = model), linewidth = 0.8) +
  geom_point(aes(y = truth), color = "black", size = 2) +
  facet_wrap(~ disease, ncol = 1, scales = "free_y") +
  labs(title = "Rolling-origin nowcasts vs. truth",
       subtitle = "Points = observed, lines = posterior median, ribbon = 50% CI",
       x = NULL, y = "Count") +
  theme_minimal()
```

## Error over time

```{r error-over-time, fig.width=10, fig.height=5}
results %>%
  mutate(abs_error = abs(truth - median)) %>%
  ggplot(aes(x = target_date, y = abs_error, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ disease, scales = "free_y") +
  labs(title = "Absolute error by week", x = NULL, y = "|truth - median|") +
  theme_minimal()
```

## Seasonal phase analysis

Does one model perform better during the **rising** phase of an outbreak vs.
the **peak** vs. the **falling** phase? We classify each target week by whether
the truth is increasing, near peak, or decreasing relative to a local window.

```{r seasonal-phase, fig.width=10, fig.height=6}
# Classify each (disease, target_date) into a phase based on the
# change in truth over a ±2-week window
phase_labels <- results %>%
  filter(model == model[1]) %>%   # one row per (disease, target_date)
  select(disease, target_date, truth) %>%
  group_by(disease) %>%
  arrange(target_date) %>%
  mutate(
    truth_lag2  = lag(truth, 2),
    truth_lead2 = lead(truth, 2),
    delta_back  = truth - truth_lag2,
    delta_fwd   = truth_lead2 - truth,
    phase = case_when(
      is.na(delta_back) | is.na(delta_fwd) ~ "edge",
      delta_back > 0 & delta_fwd >= 0      ~ "rising",
      delta_back >= 0 & delta_fwd < 0      ~ "peak",
      delta_back < 0 & delta_fwd <= 0      ~ "falling",
      delta_back < 0 & delta_fwd > 0       ~ "trough",
      TRUE                                  ~ "stable"
    )
  ) %>%
  ungroup() %>%
  select(disease, target_date, phase)

results_phase <- results %>%
  left_join(phase_labels, by = c("disease", "target_date")) %>%
  filter(phase != "edge")

# CRPS by model x disease x phase
phase_metrics <- results_phase %>%
  group_by(model, disease, phase) %>%
  summarise(
    n    = n(),
    CRPS = mean(crps),
    MAE  = mean(abs(truth - median)),
    .groups = "drop"
  )

cat("=== Metrics by seasonal phase ===\n")
print(phase_metrics %>% arrange(disease, phase, CRPS))

ggplot(results_phase, aes(x = model, y = crps, fill = model)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.3) +
  facet_grid(phase ~ disease, scales = "free_y") +
  labs(title = "CRPS by disease and seasonal phase",
       subtitle = "rising = counts increasing, peak = local max, falling = counts decreasing, trough = local min",
       x = NULL, y = "CRPS") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Season-level analysis (by respiratory season)

Group by respiratory season (Oct–Sep) to see whether model rankings are
consistent across years or if one model shines in specific seasons.

```{r by-season, fig.width=10, fig.height=6}
results_season <- results %>%
  mutate(
    # Respiratory season: Oct 2024 - Sep 2025 = "2024-25"
    resp_year = if_else(month(target_date) >= 10,
                        year(target_date),
                        year(target_date) - 1),
    season = paste0(resp_year, "-", str_sub(as.character(resp_year + 1), 3, 4))
  )

season_metrics <- results_season %>%
  group_by(model, disease, season) %>%
  summarise(
    n    = n(),
    CRPS = mean(crps),
    MAE  = mean(abs(truth - median)),
    .groups = "drop"
  )

cat("=== Metrics by respiratory season ===\n")
print(season_metrics %>% arrange(disease, season, CRPS))

ggplot(season_metrics, aes(x = model, y = CRPS, fill = model)) +
  geom_col(alpha = 0.7, position = "dodge") +
  facet_grid(season ~ disease, scales = "free_y") +
  labs(title = "Mean CRPS by model, disease, and respiratory season",
       x = NULL, y = "Mean CRPS") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Relative skill score (vs. base model)

A **skill score** measures how much better (or worse) each model is relative
to the base model: `skill = 1 - CRPS_model / CRPS_base`. Positive = better
than base, negative = worse.

```{r skill-score, fig.width=9, fig.height=6}
# Per-week skill relative to base_kinsa
base_crps <- results %>%
  filter(model == "base_kinsa") %>%
  select(disease, target_date, crps_base = crps)

skill_results <- results %>%
  filter(model != "base_kinsa") %>%
  left_join(base_crps, by = c("disease", "target_date")) %>%
  mutate(skill = 1 - crps / crps_base)

# Summary skill scores
skill_summary <- skill_results %>%
  group_by(model, disease) %>%
  summarise(
    mean_skill = mean(skill),
    pct_better = mean(skill > 0) * 100,   # % of weeks this model beats base
    .groups = "drop"
  )

cat("=== Skill scores relative to base_kinsa ===\n")
cat("mean_skill > 0 = better than base on average\n")
cat("pct_better = % of weeks where model beats base\n\n")
print(skill_summary %>% arrange(disease, desc(mean_skill)))

ggplot(skill_results, aes(x = target_date, y = skill, color = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap(~ disease, ncol = 1, scales = "free_y") +
  labs(title = "CRPS skill score relative to base model",
       subtitle = "Above 0 = model outperforms base that week",
       x = NULL, y = "Skill (1 - CRPS_model / CRPS_base)") +
  theme_minimal()
```

## Paired Wilcoxon test: is any model significantly better?

A formal paired test on per-week CRPS. Each week is a matched pair
(same truth, same data vintage) — so a paired test is appropriate.

```{r paired-tests}
cat("=== Paired Wilcoxon signed-rank tests (CRPS) ===\n")
cat("H0: model and base_kinsa have the same CRPS distribution\n")
cat("H1: they differ (two-sided)\n\n")

base_scores <- results %>%
  filter(model == "base_kinsa") %>%
  select(disease, target_date, crps_base = crps)

for (mname in setdiff(unique(results$model), "base_kinsa")) {
  cat("--- ", mname, " vs base_kinsa ---\n")
  model_scores <- results %>%
    filter(model == mname) %>%
    select(disease, target_date, crps_model = crps) %>%
    inner_join(base_scores, by = c("disease", "target_date"))

  for (d in unique(model_scores$disease)) {
    pair <- model_scores %>% filter(disease == d)
    wt <- wilcox.test(pair$crps_model, pair$crps_base, paired = TRUE)
    direction <- if (mean(pair$crps_model) < mean(pair$crps_base)) "BETTER" else "WORSE"
    cat(sprintf("  %s: p=%.4f  mean diff=%.1f  (%s)\n",
                d, wt$p.value,
                mean(pair$crps_model - pair$crps_base),
                direction))
  }
  cat("\n")
}
```

## Calibration check

Are the prediction intervals well-calibrated? For a perfectly calibrated
model, X% of observations should fall within the X% interval.

```{r calibration, fig.width=8, fig.height=5}
calibration <- results %>%
  group_by(model, disease) %>%
  summarise(
    `50% CI` = mean(truth >= q25  & truth <= q75)  * 100,
    `95% CI` = mean(truth >= q025 & truth <= q975) * 100,
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`50% CI`, `95% CI`),
               names_to = "interval", values_to = "coverage")

ggplot(calibration, aes(x = model, y = coverage, fill = model)) +
  geom_col(alpha = 0.7) +
  geom_hline(aes(yintercept = as.numeric(str_extract(interval, "\\d+"))),
             linetype = "dashed", color = "gray30") +
  facet_grid(interval ~ disease, scales = "free_y") +
  labs(title = "Empirical coverage vs. nominal",
       subtitle = "Dashed line = target coverage",
       x = NULL, y = "Coverage (%)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Paper-Ready Tables

The tables below use `kableExtra` for LaTeX/HTML formatting. If you're
writing in LaTeX, use `format = "latex"` below; for an HTML knit keep as-is.

```{r table-setup, message=FALSE}
library(kableExtra)

# Nicer model labels for the paper
model_labels <- c(
  base_kinsa = "Base + Kinsa",
  m1_kinsa   = "M1 (Diseases → Kinsa)",
  m2_kinsa   = "M2 (Kinsa → Diseases)",
  m3_kinsa   = "M3 MVN + Kinsa",
  base_ili   = "Base + ILI",
  m1_ili     = "M1 (Diseases → ILI)",
  m2_ili     = "M2 (ILI → Diseases)",
  m3_ili     = "M3 MVN + ILI"
)

results_labeled <- results %>%
  mutate(
    model = recode(model, !!!model_labels),
    disease = recode(disease, flu = "Influenza", rsv = "RSV", covid = "COVID-19")
  )
```

### Table 1: Overall nowcast accuracy by model and disease

This is the main results table. Rows = models, columns grouped by disease,
cells = MAE / CRPS / 95% coverage. Bold the best value per column.

```{r table-1-overall}
tbl1 <- results_labeled %>%
  group_by(model, disease) %>%
  summarise(
    MAE       = mean(abs(truth - median)),
    CRPS      = mean(crps),
    `95% Cov` = mean(truth >= q025 & truth <= q975) * 100,
    .groups = "drop"
  ) %>%
  mutate(across(c(MAE, CRPS), ~ round(.x, 1)),
         `95% Cov` = round(`95% Cov`, 0))

# Pivot wider so diseases are column groups
tbl1_wide <- tbl1 %>%
  pivot_wider(
    names_from = disease,
    values_from = c(MAE, CRPS, `95% Cov`),
    names_glue = "{disease}_{.value}"
  ) %>%
  select(model,
         starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

# Bold the best (lowest MAE/CRPS, closest-to-95 coverage) per column
bold_best <- function(x, target = NA) {
  if (is.na(target)) {
    best <- which.min(x)
  } else {
    best <- which.min(abs(x - target))
  }
  out <- as.character(x)
  out[best] <- cell_spec(out[best], bold = TRUE)
  out
}

tbl1_fmt <- tbl1_wide %>%
  mutate(
    across(contains("MAE"),     ~ bold_best(.x)),
    across(contains("CRPS"),    ~ bold_best(.x)),
    across(contains("Cov"),     ~ bold_best(.x, target = 95))
  )

kbl(tbl1_fmt, escape = FALSE, align = "lrrrrrrrrr",
    caption = "Table 1: Nowcast performance by model and disease (100-week rolling-origin backtest). Bold = best per column.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 1,
                     "Influenza" = 3, "RSV" = 3, "COVID-19" = 3))
```

### Table 2: CRPS skill scores relative to the base model

Shows how much each structured model improves (or worsens) relative to the
independent base. Includes % of weeks that beat base and a Wilcoxon p-value.

```{r table-2-skill}
base_crps_labeled <- results_labeled %>%
  filter(model == "Base (Independent)") %>%
  select(disease, target_date, crps_base = crps)

skill_tbl <- results_labeled %>%
  filter(model != "Base (Independent)") %>%
  left_join(base_crps_labeled, by = c("disease", "target_date")) %>%
  mutate(skill = 1 - crps / crps_base) %>%
  group_by(model, disease) %>%
  summarise(
    `Mean Skill` = round(mean(skill), 3),
    `% Weeks Better` = round(mean(skill > 0) * 100, 0),
    `p-value` = {
      wt <- wilcox.test(crps, crps_base, paired = TRUE)
      format.pval(wt$p.value, digits = 3, eps = 0.001)
    },
    .groups = "drop"
  )

skill_tbl_wide <- skill_tbl %>%
  pivot_wider(
    names_from = disease,
    values_from = c(`Mean Skill`, `% Weeks Better`, `p-value`),
    names_glue = "{disease}_{.value}"
  ) %>%
  select(model,
         starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

kbl(skill_tbl_wide, escape = FALSE, align = "lrrrrrrrrr",
    caption = "Table 2: CRPS skill score relative to Base model. Skill > 0 = improvement. Wilcoxon paired signed-rank test.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 1,
                     "Influenza" = 3, "RSV" = 3, "COVID-19" = 3))
```

### Table 3: Performance by seasonal phase

Shows which models excel during the rising, peak, falling, and trough
phases of each disease's cycle.

```{r table-3-phase}
phase_labels_labeled <- phase_labels %>%
  mutate(disease = recode(disease, flu = "Influenza", rsv = "RSV", covid = "COVID-19"))

phase_tbl <- results_labeled %>%
  left_join(phase_labels_labeled, by = c("disease", "target_date")) %>%
  filter(!is.na(phase), phase != "edge") %>%
  group_by(model, disease, phase) %>%
  summarise(
    n    = n(),
    CRPS = round(mean(crps), 1),
    .groups = "drop"
  )

# Pivot: rows = model x phase, columns = disease
phase_tbl_wide <- phase_tbl %>%
  select(-n) %>%
  pivot_wider(names_from = disease, values_from = CRPS) %>%
  arrange(phase, model)

kbl(phase_tbl_wide, align = "llrrr",
    caption = "Table 3: Mean CRPS by model and seasonal phase.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  pack_rows(index = table(phase_tbl_wide$phase))
```

### Table 4: Performance by respiratory season

```{r table-4-season}
results_season_labeled <- results_labeled %>%
  mutate(
    resp_year = if_else(month(target_date) >= 10,
                        year(target_date),
                        year(target_date) - 1),
    season = paste0(resp_year, "-", str_sub(as.character(resp_year + 1), 3, 4))
  )

season_tbl <- results_season_labeled %>%
  group_by(model, disease, season) %>%
  summarise(
    n    = n(),
    CRPS = round(mean(crps), 1),
    MAE  = round(mean(abs(truth - median)), 1),
    .groups = "drop"
  )

season_tbl_wide <- season_tbl %>%
  select(-n) %>%
  pivot_wider(
    names_from = disease,
    values_from = c(CRPS, MAE),
    names_glue = "{disease}_{.value}"
  ) %>%
  arrange(season, model) %>%
  select(season, model,
         starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

kbl(season_tbl_wide, align = "llrrrrrr",
    caption = "Table 4: Mean CRPS and MAE by model and respiratory season (Oct-Sep).") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 2,
                     "Influenza" = 2, "RSV" = 2, "COVID-19" = 2)) %>%
  pack_rows(index = table(season_tbl_wide$season))
```
