---
title: "Base Nowcast Model"
output: html_document
date: "2026-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rjags)
library(tidyverse)
library(coda)
library(tidycensus)
library(HDInterval)
library(patchwork)
```

```{r}
#census_api_key("81e4817f989c60bf268764565acf7fb7fbe0e2cf", install = TRUE)
```

## Load in the Data
```{r load_data}
data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz")
ili_data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/delphi_ili_fluview/standard/data.csv.gz")


# filtering the data to state and using "total" as the age demographic

fips <- "26" # Michigan

data <- data %>%
  filter(geography == fips , age == "Total")

# Filter ILI data for state
ili_data <- ili_data %>%
  filter(geography == fips) %>%
  select(time, ili_count = delphi_fluview_num_ili) %>%
  arrange(time)

state_pre <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2019,
  time_series = TRUE
) |>
  filter(variable == "POP") |>
  mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |>
  arrange(year)|>
  pull(value)

state_pop <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2025,  # Most recent available
  time_series = TRUE
) |>
  filter(variable == "POPESTIMATE", year %in% c(2020:2025)) %>%
  arrange(year) %>%
  pull(value)


state_pop_estimates <- tibble(
  year = 2018:2026,
  population = c(
    # State census population estimates
    state_pre[1], #2018
    state_pre[2], #2019
    state_pop[1],   #2020
    state_pop[2],   #2021
    state_pop[3],   #2022
    state_pop[4],   #2023
    state_pop[5],   #2024
    state_pop[6],   #2025
    state_pop[6]   #2026 - using 2025 estimate for 2026
  )
)

```

## Merging Population Estimates with Data
```{r population_merge}
data_with_pop <- data %>%
  mutate(year = year(time)) %>%
  left_join(state_pop_estimates, by = "year")

data_with_pop <- data_with_pop %>%
  mutate(
    covid_count = round(rate_covid * population / 100000),
    flu_count = round(rate_flu * population / 100000),
    rsv_count = round(rate_rsv * population / 100000)
  )

# Merge ILI data with the main dataset
data_with_pop <- data_with_pop %>%
  left_join(ili_data, by = "time") %>%
  mutate(ili_count = replace_na(ili_count, 0))
```

## JAGS Model with NA-Based Evaluation

The model uses NA values for held-out observations, which JAGS will automatically predict.

```{r jags-model}
model_string <- "
model {
  # Likelihood for all time points
  # NA values will be automatically predicted by JAGS
  for(i in 1:tmax) {
    flu[i] ~ dpois(lambda1[i])
    rsv[i] ~ dpois(lambda2[i])
    covid[i] ~ dpois(lambda3[i])
    ili[i] ~ dpois(lambda4[i])
  }

  # Log-linear model for rates
  for(i in 1:tmax) {
    lambda1[i] <- exp(phi[i,1])
    lambda2[i] <- exp(phi[i,2])
    lambda3[i] <- exp(phi[i,3])
    lambda4[i] <- exp(phi[i,4])
  }

  # Priors - AR(1) process for each disease
  for(k in 1:4) {
    mu0[k] ~ dnorm(0, 1e-4)
    prec.phi[k] ~ dgamma(0.01, 0.01)
    rho1[k] ~ dunif(0, 1)

    # First time point - stationary distribution
    # Stationary precision = prec.phi * (1 - rho^2)
    phi[1,k] ~ dnorm(mu0[k], prec.phi[k] * (1 - pow(rho1[k], 2)))

    # AR(1) evolution for subsequent time points
    for(i in 2:tmax) {
      phi[i,k] ~ dnorm(mu0[k] + rho1[k] * (phi[i-1,k] - mu0[k]), prec.phi[k])
    }
  }
}
"

writeLines(model_string, "base_model.jags")
```

## Prepare JAGS Data

```{r jags-data}
# Number of weeks to hold out for evaluation
n_holdout <- 2

tmax <- nrow(data_with_pop)

cat("Total observations:", tmax, "\n")
cat("Training on:", tmax - n_holdout, "observations\n")
cat("Evaluating on last:", n_holdout, "observations\n")

# Store the actual values for comparison BEFORE setting to NA
actuals_holdout <- data_with_pop %>%
  tail(n_holdout) %>%
  select(time, flu_count, rsv_count, covid_count, ili_count)

cat("\nHeld-out actual values:\n")
print(actuals_holdout)

# Create data vectors with NA for held-out observations
# JAGS will automatically predict these NA values
flu_data <- data_with_pop$flu_count
rsv_data <- data_with_pop$rsv_count
covid_data <- data_with_pop$covid_count
ili_data_vec <- data_with_pop$ili_count

# Set last n_holdout observations to NA
flu_data[(tmax - n_holdout + 1):tmax] <- NA
rsv_data[(tmax - n_holdout + 1):tmax] <- NA
covid_data[(tmax - n_holdout + 1):tmax] <- NA
ili_data_vec[(tmax - n_holdout + 1):tmax] <- NA

# Prepare data list for JAGS
jags_data <- list(
  flu = flu_data,
  rsv = rsv_data,
  covid = covid_data,
  ili = ili_data_vec,
  tmax = tmax
)

cat("\nLast few values passed to JAGS (NA = to be predicted):\n")
cat("Flu:", tail(jags_data$flu, 5), "\n")
cat("RSV:", tail(jags_data$rsv, 5), "\n")
cat("COVID:", tail(jags_data$covid, 5), "\n")
cat("ILI:", tail(jags_data$ili, 5), "\n")
```

## Run JAGS Model

```{r run-jags, message=FALSE, warning=FALSE}
# Parameters to monitor
# flu, rsv, covid, ili will include predicted values for the NA observations
params <- c("mu0", "rho1", "prec.phi", "lambda1", "lambda2", "lambda3", "lambda4",
            "flu", "rsv", "covid", "ili")

# Set seeds for reproducibility
inits <- list(
  list(".RNG.seed" = 123, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 456, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 789, ".RNG.name" = "base::Wichmann-Hill")
)

# Initialize the model
jags_model <- jags.model(
  file = "base_model.jags",
  data = jags_data,
  n.chains = 3,
  n.adapt = 1000,
  inits = inits
)

# Burn-in
update(jags_model, n.iter = 10000)

samples <- coda.samples(
  model = jags_model,
  variable.names = params,
  n.iter = 20000,
  thin = 10
)
```

## Model Diagnostics

```{r diagnostics}
# Summary of key parameters
summary(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                    "rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]")])

# Trace plots for AR(1) parameters
plot(samples[, c("rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]")])

# Gelman-Rubin diagnostic (should be close to 1)
gelman.diag(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                        "rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]")])
```


## Extract Predictions and Evaluate

```{r extract-predictions, warning=FALSE}
# Combine chains
samples_combined <- do.call(rbind, samples)

# Extract predictions for held-out weeks
eval_results <- tibble(
  week = 1:n_holdout,
  time = actuals_holdout$time,

  # Actual values
  actual_flu = actuals_holdout$flu_count,
  actual_rsv = actuals_holdout$rsv_count,
  actual_covid = actuals_holdout$covid_count,
  actual_ili = actuals_holdout$ili_count
)

# Add predictions for each held-out week
for (h in 1:n_holdout) {
  idx <- (tmax - n_holdout) + h

  # Extract posterior samples for predictions (JAGS predicted these NA values)
  flu_samples <- samples_combined[, paste0("flu[", idx, "]")]
  rsv_samples <- samples_combined[, paste0("rsv[", idx, "]")]
  covid_samples <- samples_combined[, paste0("covid[", idx, "]")]
  ili_samples <- samples_combined[, paste0("ili[", idx, "]")]

  # Mean estimates
  eval_results$pred_flu_mean[h] <- mean(flu_samples)
  eval_results$pred_rsv_mean[h] <- mean(rsv_samples)
  eval_results$pred_covid_mean[h] <- mean(covid_samples)
  eval_results$pred_ili_mean[h] <- mean(ili_samples)

  # Median estimates
  eval_results$pred_flu_median[h] <- median(flu_samples)
  eval_results$pred_rsv_median[h] <- median(rsv_samples)
  eval_results$pred_covid_median[h] <- median(covid_samples)
  eval_results$pred_ili_median[h] <- median(ili_samples)

  # 95% HDI (Highest Density Interval) using HDInterval package
  flu_hdi <- hdi(flu_samples, credMass = 0.95)
  rsv_hdi <- hdi(rsv_samples, credMass = 0.95)
  covid_hdi <- hdi(covid_samples, credMass = 0.95)
  ili_hdi <- hdi(ili_samples, credMass = 0.95)

  eval_results$pred_flu_hdi_lower[h] <- flu_hdi[1]
  eval_results$pred_flu_hdi_upper[h] <- flu_hdi[2]
  eval_results$pred_rsv_hdi_lower[h] <- rsv_hdi[1]
  eval_results$pred_rsv_hdi_upper[h] <- rsv_hdi[2]
  eval_results$pred_covid_hdi_lower[h] <- covid_hdi[1]
  eval_results$pred_covid_hdi_upper[h] <- covid_hdi[2]
  eval_results$pred_ili_hdi_lower[h] <- ili_hdi[1]
  eval_results$pred_ili_hdi_upper[h] <- ili_hdi[2]
}

# Calculate errors and coverage
eval_results <- eval_results %>%
  mutate(
    # Absolute errors using mean
    abs_error_flu_mean = abs(actual_flu - pred_flu_mean),
    abs_error_rsv_mean = abs(actual_rsv - pred_rsv_mean),
    abs_error_covid_mean = abs(actual_covid - pred_covid_mean),
    abs_error_ili_mean = abs(actual_ili - pred_ili_mean),

    # Absolute errors using median
    abs_error_flu_median = abs(actual_flu - pred_flu_median),
    abs_error_rsv_median = abs(actual_rsv - pred_rsv_median),
    abs_error_covid_median = abs(actual_covid - pred_covid_median),
    abs_error_ili_median = abs(actual_ili - pred_ili_median),

    # Coverage: is actual within 95% HDI?
    covered_flu = actual_flu >= pred_flu_hdi_lower & actual_flu <= pred_flu_hdi_upper,
    covered_rsv = actual_rsv >= pred_rsv_hdi_lower & actual_rsv <= pred_rsv_hdi_upper,
    covered_covid = actual_covid >= pred_covid_hdi_lower & actual_covid <= pred_covid_hdi_upper,
    covered_ili = actual_ili >= pred_ili_hdi_lower & actual_ili <= pred_ili_hdi_upper
  )
```

## Prediction Results

```{r prediction-results}
cat("=== FLU Predictions ===\n")
eval_results %>%
  select(time, actual_flu, pred_flu_mean, pred_flu_median,
         pred_flu_hdi_lower, pred_flu_hdi_upper, covered_flu) %>%
  print()

cat("\n=== RSV Predictions ===\n")
eval_results %>%
  select(time, actual_rsv, pred_rsv_mean, pred_rsv_median,
         pred_rsv_hdi_lower, pred_rsv_hdi_upper, covered_rsv) %>%
  print()

cat("\n=== COVID Predictions ===\n")
eval_results %>%
  select(time, actual_covid, pred_covid_mean, pred_covid_median,
         pred_covid_hdi_lower, pred_covid_hdi_upper, covered_covid) %>%
  print()

cat("\n=== ILI Predictions ===\n")
eval_results %>%
  select(time, actual_ili, pred_ili_mean, pred_ili_median,
         pred_ili_hdi_lower, pred_ili_hdi_upper, covered_ili) %>%
  print()
```

## Mean vs Median Comparison

```{r mean-vs-median}
cat("=== Absolute Errors: Mean vs Median Point Estimates ===\n\n")

comparison <- tibble(
  Disease = c("Flu", "RSV", "COVID", "ILI"),
  `MAE (Mean)` = c(
    mean(eval_results$abs_error_flu_mean),
    mean(eval_results$abs_error_rsv_mean),
    mean(eval_results$abs_error_covid_mean),
    mean(eval_results$abs_error_ili_mean)
  ),
  `MAE (Median)` = c(
    mean(eval_results$abs_error_flu_median),
    mean(eval_results$abs_error_rsv_median),
    mean(eval_results$abs_error_covid_median),
    mean(eval_results$abs_error_ili_median)
  )
) %>%
  mutate(
    `Better Estimate` = ifelse(`MAE (Mean)` < `MAE (Median)`, "Mean", "Median")
  )

print(comparison)
```

## Evaluation Metrics

```{r evaluation-metrics}
# Calculate summary metrics
metrics <- tibble(
  Disease = c("Flu", "RSV", "COVID", "ILI"),
  `MAE (Median)` = c(
    mean(eval_results$abs_error_flu_median),
    mean(eval_results$abs_error_rsv_median),
    mean(eval_results$abs_error_covid_median),
    mean(eval_results$abs_error_ili_median)
  ),
  `RMSE (Median)` = c(
    sqrt(mean(eval_results$abs_error_flu_median^2)),
    sqrt(mean(eval_results$abs_error_rsv_median^2)),
    sqrt(mean(eval_results$abs_error_covid_median^2)),
    sqrt(mean(eval_results$abs_error_ili_median^2))
  ),
  `95% HDI Coverage` = c(
    mean(eval_results$covered_flu),
    mean(eval_results$covered_rsv),
    mean(eval_results$covered_covid),
    mean(eval_results$covered_ili)
  )
)

cat("=== Evaluation Metrics ===\n")
cat("MAE = Mean Absolute Error\n")
cat("RMSE = Root Mean Squared Error\n")
cat("Coverage = Proportion of actuals within 95% HDI (target: 0.95)\n\n")
print(metrics)
```

## Extract Fitted Values for Training Period

```{r fitted-values}
# Extract lambda (fitted rates) for all time points
lambda1_cols <- grep("^lambda1\\[", colnames(samples_combined))
lambda2_cols <- grep("^lambda2\\[", colnames(samples_combined))
lambda3_cols <- grep("^lambda3\\[", colnames(samples_combined))
lambda4_cols <- grep("^lambda4\\[", colnames(samples_combined))

# Posterior medians (all time points)
fitted_flu <- apply(samples_combined[, lambda1_cols], 2, median)
fitted_rsv <- apply(samples_combined[, lambda2_cols], 2, median)
fitted_covid <- apply(samples_combined[, lambda3_cols], 2, median)
fitted_ili <- apply(samples_combined[, lambda4_cols], 2, median)

# 95% HDI for fitted values
fitted_flu_hdi <- apply(samples_combined[, lambda1_cols], 2, function(x) hdi(x, 0.95))
fitted_rsv_hdi <- apply(samples_combined[, lambda2_cols], 2, function(x) hdi(x, 0.95))
fitted_covid_hdi <- apply(samples_combined[, lambda3_cols], 2, function(x) hdi(x, 0.95))
fitted_ili_hdi <- apply(samples_combined[, lambda4_cols], 2, function(x) hdi(x, 0.95))

# Create results dataframe for full time series
results <- data_with_pop %>%
  mutate(
    fitted_flu = fitted_flu,
    fitted_flu_lower = fitted_flu_hdi[1, ],
    fitted_flu_upper = fitted_flu_hdi[2, ],
    fitted_rsv = fitted_rsv,
    fitted_rsv_lower = fitted_rsv_hdi[1, ],
    fitted_rsv_upper = fitted_rsv_hdi[2, ],
    fitted_covid = fitted_covid,
    fitted_covid_lower = fitted_covid_hdi[1, ],
    fitted_covid_upper = fitted_covid_hdi[2, ],
    fitted_ili = fitted_ili,
    fitted_ili_lower = fitted_ili_hdi[1, ],
    fitted_ili_upper = fitted_ili_hdi[2, ],
    period = ifelse(row_number() <= (tmax - n_holdout), "Training", "Evaluation")
  )
```

## Visualization

```{r plots, fig.width=10, fig.height=13}
# Train/eval split point
split_time <- results$time[tmax - n_holdout]

# Flu
p1 <- ggplot(results, aes(x = time)) +
  geom_ribbon(aes(ymin = fitted_flu_lower, ymax = fitted_flu_upper),
              fill = "blue", alpha = 0.2) +
  geom_line(aes(y = fitted_flu), color = "blue", linewidth = 1) +
  geom_point(aes(y = flu_count, shape = period), alpha = 0.7, size = 2) +
  geom_vline(xintercept = split_time, linetype = "dashed", color = "gray40") +
  scale_shape_manual(values = c("Training" = 16, "Evaluation" = 17)) +
  labs(title = "Flu: Observed vs Fitted (Median + 95% HDI)",
       subtitle = "Dashed line = train/eval split, Triangle = held-out",
       y = "Count", x = "Time") +
  theme_minimal()

# RSV
p2 <- ggplot(results, aes(x = time)) +
  geom_ribbon(aes(ymin = fitted_rsv_lower, ymax = fitted_rsv_upper),
              fill = "green", alpha = 0.2) +
  geom_line(aes(y = fitted_rsv), color = "darkgreen", linewidth = 1) +
  geom_point(aes(y = rsv_count, shape = period), alpha = 0.7, size = 2) +
  geom_vline(xintercept = split_time, linetype = "dashed", color = "gray40") +
  scale_shape_manual(values = c("Training" = 16, "Evaluation" = 17)) +
  labs(title = "RSV: Observed vs Fitted (Median + 95% HDI)", y = "Count", x = "Time") +
  theme_minimal()

# COVID
p3 <- ggplot(results, aes(x = time)) +
  geom_ribbon(aes(ymin = fitted_covid_lower, ymax = fitted_covid_upper),
              fill = "red", alpha = 0.2) +
  geom_line(aes(y = fitted_covid), color = "darkred", linewidth = 1) +
  geom_point(aes(y = covid_count, shape = period), alpha = 0.7, size = 2) +
  geom_vline(xintercept = split_time, linetype = "dashed", color = "gray40") +
  scale_shape_manual(values = c("Training" = 16, "Evaluation" = 17)) +
  labs(title = "COVID: Observed vs Fitted (Median + 95% HDI)", y = "Count", x = "Time") +
  theme_minimal()

# ILI
p4 <- ggplot(results, aes(x = time)) +
  geom_ribbon(aes(ymin = fitted_ili_lower, ymax = fitted_ili_upper),
              fill = "orange", alpha = 0.2) +
  geom_line(aes(y = fitted_ili), color = "darkorange", linewidth = 1) +
  geom_point(aes(y = ili_count, shape = period), alpha = 0.7, size = 2) +
  geom_vline(xintercept = split_time, linetype = "dashed", color = "gray40") +
  scale_shape_manual(values = c("Training" = 16, "Evaluation" = 17)) +
  labs(title = "ILI: Observed vs Fitted (Median + 95% HDI)", y = "Count", x = "Time") +
  theme_minimal()

p1 / p2 / p3 / p4
```

## Model Summary

```{r summary}
# AR(1) parameters for each disease
cat("=== AR(1) Autocorrelation Parameters (rho) ===\n")
cat("Flu:   Median =", round(median(samples_combined[, "rho1[1]"]), 3),
    " | Mean =", round(mean(samples_combined[, "rho1[1]"]), 3), "\n")
cat("RSV:   Median =", round(median(samples_combined[, "rho1[2]"]), 3),
    " | Mean =", round(mean(samples_combined[, "rho1[2]"]), 3), "\n")
cat("COVID: Median =", round(median(samples_combined[, "rho1[3]"]), 3),
    " | Mean =", round(mean(samples_combined[, "rho1[3]"]), 3), "\n")
cat("ILI:   Median =", round(median(samples_combined[, "rho1[4]"]), 3),
    " | Mean =", round(mean(samples_combined[, "rho1[4]"]), 3), "\n")

# Mean levels
cat("\n=== Mean Log-Rate Parameters (mu0) ===\n")
cat("Flu:   Median =", round(median(samples_combined[, "mu0[1]"]), 3),
    " | Mean =", round(mean(samples_combined[, "mu0[1]"]), 3), "\n")
cat("RSV:   Median =", round(median(samples_combined[, "mu0[2]"]), 3),
    " | Mean =", round(mean(samples_combined[, "mu0[2]"]), 3), "\n")
cat("COVID: Median =", round(median(samples_combined[, "mu0[3]"]), 3),
    " | Mean =", round(mean(samples_combined[, "mu0[3]"]), 3), "\n")
cat("ILI:   Median =", round(median(samples_combined[, "mu0[4]"]), 3),
    " | Mean =", round(mean(samples_combined[, "mu0[4]"]), 3), "\n")

cat("\n=== Evaluation Summary ===\n")
cat("Held out last", n_holdout, "weeks for evaluation\n")
print(metrics)
```
