---
title: "Base Nowcast Model (Kinsa)"
output: html_document
date: "2026-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rjags)
library(tidyverse)
library(coda)
library(tidycensus)
library(HDInterval)
library(patchwork)
library(scoringRules)
```

```{r}
#census_api_key("81e4817f989c60bf268764565acf7fb7fbe0e2cf", install = TRUE)
```

## Load in the Data

This mirrors `Base_Model.Rmd` but replaces the ILI FluView signal with Kinsa
networked-thermometer data (`percent_ill`, aggregated weekly). Because Kinsa
is updated daily while RESP-NET has a one-week reporting lag, this file
simulates the scenario: Kinsa is real-time (no holdout), RESP-NET signals
(flu/rsv/covid) are held out by one week.

```{r load_data}
# -------- FILL IN: local path to your Kinsa CSV --------
kinsa_path <- "/Users/tobysalmon/Desktop/kinsa_michigan_weekly.csv"
# -------------------------------------------------------

# Name of the state inside Kinsa's `region_name` column (must match exactly)
kinsa_state <- "Michigan"

data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz")
kinsa_raw <- read_csv(kinsa_path)

# filtering the data to state and using "total" as the age demographic
fips <- "26" # Michigan

data <- data %>%
  filter(geography == fips , age == "Total")

# Filter Kinsa data for state and put it on the same weekly grid
kinsa_data <- kinsa_raw %>%
  filter(region_name == kinsa_state) %>%
  mutate(time = as.Date(time)) %>%
  select(time, percent_ill) %>%
  arrange(time)

state_pre <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2019,
  time_series = TRUE
) |>
  filter(variable == "POP") |>
  mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |>
  arrange(year)|>
  pull(value)

state_pop <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2025,
  time_series = TRUE
) |>
  filter(variable == "POPESTIMATE", year %in% c(2020:2025)) %>%
  arrange(year) %>%
  pull(value)

state_pop_estimates <- tibble(
  year = 2018:2026,
  population = c(
    state_pre[1], #2018
    state_pre[2], #2019
    state_pop[1], #2020
    state_pop[2], #2021
    state_pop[3], #2022
    state_pop[4], #2023
    state_pop[5], #2024
    state_pop[6], #2025
    state_pop[6]  #2026 - using 2025 estimate
  )
)
```

## Merging Population Estimates with Data

```{r population_merge}
data_with_pop <- data %>%
  mutate(year = year(time)) %>%
  left_join(state_pop_estimates, by = "year")

data_with_pop <- data_with_pop %>%
  mutate(
    covid_count = round(rate_covid * population / 100000),
    flu_count   = round(rate_flu   * population / 100000),
    rsv_count   = round(rate_rsv   * population / 100000)
  )

# Merge Kinsa into the main dataset. Inner-join on time so the panel is
# restricted to weeks where both sources exist. This also guarantees
# percent_ill is present (lognormal requires strictly positive values).
data_with_pop <- data_with_pop %>%
  inner_join(kinsa_data, by = "time") %>%
  arrange(time)

# Sanity check: lognormal needs percent_ill > 0
stopifnot(all(data_with_pop$percent_ill > 0))
```

## JAGS Model with NA-Based Evaluation

Same structure as the base model, but the fourth series is Kinsa `percent_ill`
modeled as lognormal: `kinsa[i] ~ dlnorm(phi[i,4], tau_kinsa)`. There is no
exposure offset because Kinsa is already a proportion.

```{r jags-model}
model_string <- "
model {
  for(i in 1:tmax) {
    flu[i]   ~ dnegbin(p1[i], r[1])
    rsv[i]   ~ dpois(lambda2[i])
    covid[i] ~ dnegbin(p3[i], r[3])
    kinsa[i] ~ dlnorm(phi[i,4], tau_kinsa)

    p1[i] <- r[1] / (r[1] + lambda1[i])
    p3[i] <- r[3] / (r[3] + lambda3[i])

    lambda1[i] <- exp(phi[i,1])
    lambda2[i] <- exp(phi[i,2])
    lambda3[i] <- exp(phi[i,3])
    # lambda4 is the median of the lognormal (exp of meanlog)
    lambda4[i] <- exp(phi[i,4])
  }

  # Seasonal AR(1) priors - same structure for all 4 series
  for(k in 1:4) {
    mu0[k] ~ dnorm(0, 1e-4)
    beta_sin[k] ~ dnorm(0, 1e-4)
    beta_cos[k] ~ dnorm(0, 1e-4)
    prec.phi[k] ~ dgamma(0.01, 0.01)
    rho1[k] ~ dunif(0, 1)

    mu_t[1,k] <- mu0[k] + beta_sin[k] * sin52[1] + beta_cos[k] * cos52[1]
    phi[1,k] ~ dnorm(mu_t[1,k], prec.phi[k] * (1 - pow(rho1[k], 2)))

    for(i in 2:tmax) {
      mu_t[i,k] <- mu0[k] + beta_sin[k] * sin52[i] + beta_cos[k] * cos52[i]
      phi[i,k] ~ dnorm(mu_t[i,k] + rho1[k] * (phi[i-1,k] - mu_t[i-1,k]),
                        prec.phi[k])
    }
  }

  # Overdispersion priors for NegBin diseases (flu, covid). RSV is Poisson.
  r[1] ~ dgamma(0.1, 0.1)
  r[3] ~ dgamma(0.1, 0.1)

  # Log-scale precision for the Kinsa lognormal
  tau_kinsa ~ dgamma(0.01, 0.01)
}
"

writeLines(model_string, "base_model_kinsa.jags")
```

## Prepare JAGS Data

```{r jags-data}
# Holdout: respnet has a 1-week lag, Kinsa is real-time (no holdout)
n_holdout_specific <- 1
n_holdout_kinsa <- 0

tmax <- nrow(data_with_pop)

# Create harmonic vectors for seasonality (52-week cycle)
t_vec <- 1:tmax
sin52 <- sin(2 * pi * t_vec / 52)
cos52 <- cos(2 * pi * t_vec / 52)

cat("Total observations:", tmax, "\n")
cat("Flu/RSV/COVID: training on", tmax - n_holdout_specific, "| evaluating last", n_holdout_specific, "\n")
cat("Kinsa: fully observed (no holdout)\n")

# Store actual values for comparison BEFORE setting to NA
actuals_holdout_specific <- data_with_pop %>%
  tail(n_holdout_specific) %>%
  select(time, flu_count, rsv_count, covid_count)

cat("\nHeld-out actual values (specific signals):\n")
print(actuals_holdout_specific)

# Create data vectors with NA for held-out observations
flu_data   <- data_with_pop$flu_count
rsv_data   <- data_with_pop$rsv_count
covid_data <- data_with_pop$covid_count
kinsa_vec  <- data_with_pop$percent_ill  # never NA

flu_data[(tmax - n_holdout_specific + 1):tmax]   <- NA
rsv_data[(tmax - n_holdout_specific + 1):tmax]   <- NA
covid_data[(tmax - n_holdout_specific + 1):tmax] <- NA

jags_data <- list(
  flu = flu_data,
  rsv = rsv_data,
  covid = covid_data,
  kinsa = kinsa_vec,
  tmax = tmax,
  sin52 = sin52,
  cos52 = cos52
)

cat("\nLast few values passed to JAGS (NA = to be predicted):\n")
cat("Flu:",   tail(jags_data$flu, 5),   "\n")
cat("RSV:",   tail(jags_data$rsv, 5),   "\n")
cat("COVID:", tail(jags_data$covid, 5), "\n")
cat("Kinsa:", tail(jags_data$kinsa, 5), "\n")
```

## Run JAGS Model

```{r run-jags, message=FALSE, warning=FALSE}
params <- c("mu0", "rho1", "prec.phi", "r", "tau_kinsa",
            "beta_sin", "beta_cos",
            "lambda1", "lambda2", "lambda3", "lambda4",
            "flu", "rsv", "covid", "kinsa")

inits <- list(
  list(".RNG.seed" = 123, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 456, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 789, ".RNG.name" = "base::Wichmann-Hill")
)

jags_model <- jags.model(
  file = "base_model_kinsa.jags",
  data = jags_data,
  n.chains = 3,
  n.adapt = 1000,
  inits = inits
)

update(jags_model, n.iter = 10000)

samples <- coda.samples(
  model = jags_model,
  variable.names = params,
  n.iter = 20000,
  thin = 10
)
```

## Model Diagnostics

```{r diagnostics}
summary(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                    "rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]",
                    "r[1]", "r[3]", "tau_kinsa",
                    "beta_sin[1]", "beta_sin[2]", "beta_sin[3]", "beta_sin[4]",
                    "beta_cos[1]", "beta_cos[2]", "beta_cos[3]", "beta_cos[4]")])

plot(samples[, "rho1[1]"])
plot(samples[, "rho1[2]"])
plot(samples[, "rho1[3]"])
plot(samples[, "rho1[4]"])

plot(samples[, "mu0[1]"])
plot(samples[, "mu0[2]"])
plot(samples[, "mu0[3]"])
plot(samples[, "mu0[4]"])

plot(samples[, "r[1]"])
plot(samples[, "r[3]"])
plot(samples[, "tau_kinsa"])

gelman.diag(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                        "rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]",
                        "r[1]", "r[3]", "tau_kinsa",
                        "beta_sin[1]", "beta_sin[2]", "beta_sin[3]", "beta_sin[4]",
                        "beta_cos[1]", "beta_cos[2]", "beta_cos[3]", "beta_cos[4]")])
```

## Extract Predictions and Evaluate

```{r evaluation, warning=FALSE}
samples_combined <- do.call(rbind, samples)

eval_results <- tibble(
  week = 1:n_holdout_specific,
  time = actuals_holdout_specific$time,
  actual_flu   = actuals_holdout_specific$flu_count,
  actual_rsv   = actuals_holdout_specific$rsv_count,
  actual_covid = actuals_holdout_specific$covid_count
)

for (h in 1:n_holdout_specific) {
  idx <- (tmax - n_holdout_specific) + h

  flu_samples   <- samples_combined[, paste0("flu[", idx, "]")]
  rsv_samples   <- samples_combined[, paste0("rsv[", idx, "]")]
  covid_samples <- samples_combined[, paste0("covid[", idx, "]")]

  eval_results$pred_flu_median[h]   <- median(flu_samples)
  eval_results$pred_rsv_median[h]   <- median(rsv_samples)
  eval_results$pred_covid_median[h] <- median(covid_samples)

  eval_results$abs_error_flu[h]   <- abs(eval_results$actual_flu[h]   - median(flu_samples))
  eval_results$abs_error_rsv[h]   <- abs(eval_results$actual_rsv[h]   - median(rsv_samples))
  eval_results$abs_error_covid[h] <- abs(eval_results$actual_covid[h] - median(covid_samples))

  eval_results$crps_flu[h]   <- crps_sample(y = eval_results$actual_flu[h],   dat = flu_samples)
  eval_results$crps_rsv[h]   <- crps_sample(y = eval_results$actual_rsv[h],   dat = rsv_samples)
  eval_results$crps_covid[h] <- crps_sample(y = eval_results$actual_covid[h], dat = covid_samples)
}

cat("=== Per-Week Results ===\n\n")
print(eval_results)
```

```{r summary-metrics}
metrics <- tibble(
  Disease = c("Flu", "RSV", "COVID"),
  MAE = c(
    mean(eval_results$abs_error_flu),
    mean(eval_results$abs_error_rsv),
    mean(eval_results$abs_error_covid)
  ),
  CRPS = c(
    mean(eval_results$crps_flu),
    mean(eval_results$crps_rsv),
    mean(eval_results$crps_covid)
  )
)

cat("=== Summary Metrics (Base + Kinsa) ===\n")
print(metrics)
```
