---
title: "Paper Figures — Bayesian Respiratory Disease Nowcasting"
output:
  html_document:
    fig_width: 10
    fig_height: 7
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.retina = 2, dpi = 300
)
```

```{r load}
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)

# ---- Output directory for paper-ready figures/tables ----
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

# ---- Load cached backtest results ----
results <- read_csv("results/backtest_predictions.csv", show_col_types = FALSE) %>%
  mutate(as_of = as.Date(as_of), target_date = as.Date(target_date))

# ---- Labels and colors ----
model_labels <- c(
  base_kinsa = "Base + Kinsa",
  m1_kinsa   = "M1 (Diseases \u2192 Kinsa)",
  m2_kinsa   = "M2 (Kinsa \u2192 Diseases)",
  m3_kinsa   = "M3 MVN + Kinsa",
  base_ili   = "Base + ILI",
  m1_ili     = "M1 (Diseases \u2192 ILI)",
  m2_ili     = "M2 (ILI \u2192 Diseases)",
  m3_ili     = "M3 MVN + ILI"
)

disease_labels <- c(flu = "Influenza", rsv = "RSV", covid = "COVID-19")

# Colors: each model STRUCTURE gets a distinct hue.
# Kinsa = dark/saturated, ILI = lighter/muted version of the same hue.
# This makes structure immediately readable while family (Kinsa vs ILI)
# is distinguishable by saturation/lightness.
## Okabe-Ito-inspired colorblind-safe palette.
## Hues are chosen so that adjacent categories (esp. M1 vs M2) differ in
## BOTH hue and luminance -- distinguishable under deuteranopia/protanopia.
## Kinsa family = saturated, ILI family = lighter tint of same hue.
model_colors <- c(
  "Base + Kinsa"             = "#0072B2",  # deep blue         (Okabe-Ito)
  "M1 (Diseases \u2192 Kinsa)"  = "#009E73",  # bluish green      (Okabe-Ito)
  "M2 (Kinsa \u2192 Diseases)"  = "#D55E00",  # vermillion        (Okabe-Ito)
  "M3 MVN + Kinsa"           = "#6A3D9A",  # dark purple
  "Base + ILI"               = "#56B4E9",  # sky blue          (Okabe-Ito)
  "M1 (Diseases \u2192 ILI)"    = "#8FD9C0",  # light bluish green (tint)
  "M2 (ILI \u2192 Diseases)"    = "#F0B27A",  # light vermillion   (tint)
  "M3 MVN + ILI"             = "#CAB2D6"   # light purple
)
signal_colors_all <- c(
  Kinsa      = "#E69F00",   # amber/orange
  ILI        = "#CC79A7",   # pink/magenta
  Influenza  = "#0072B2",   # deep blue
  RSV        = "#009E73",   # teal green
  `COVID-19` = "#D55E00"    # vermillion red
)

results <- results %>%
  mutate(
    model   = recode(model, !!!model_labels),
    disease = recode(disease, !!!disease_labels),
    model   = factor(model, levels = model_labels)
  )

# Common theme
theme_paper <- theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title    = element_blank(),
    strip.text      = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "gray30", size = 11)
  )
```

# Descriptive Signal Analysis

```{r raw-data-load, message=FALSE, warning=FALSE}
library(tidycensus)

kinsa_path  <- "/Users/tobysalmon/Desktop/kinsa_michigan_weekly.csv"
fips        <- "26"
kinsa_state <- "Michigan"

resp_raw <- read_csv(
  "https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz",
  show_col_types = FALSE
) %>% filter(geography == fips, age == "Total")

ili_raw <- read_csv(
  "https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/delphi_ili_fluview/standard/data.csv.gz",
  show_col_types = FALSE
) %>%
  filter(geography == fips) %>%
  select(time, ili = delphi_fluview_num_ili)

kinsa_raw <- read_csv(kinsa_path, show_col_types = FALSE) %>%
  filter(region_name == kinsa_state) %>%
  mutate(time = as.Date(time)) %>%
  select(time, kinsa = percent_ill)

# Population (same vintage as Backtest.Rmd)
state_pre <- get_estimates(geography = "state", state = fips,
  product = "population", vintage = 2019, time_series = TRUE) |>
  filter(variable == "POP") |> mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |> arrange(year) |> pull(value)
state_pop <- get_estimates(geography = "state", state = fips,
  product = "population", vintage = 2025, time_series = TRUE) |>
  filter(variable == "POPESTIMATE", year %in% 2020:2025) |>
  arrange(year) |> pull(value)
pop_tbl <- tibble(year = 2018:2026,
                  population = c(state_pre, state_pop, state_pop[6]))

signal_data <- resp_raw %>%
  mutate(year = year(time)) %>%
  left_join(pop_tbl, by = "year") %>%
  mutate(
    flu   = round(rate_flu   * population / 100000),
    rsv   = round(rate_rsv   * population / 100000),
    covid = round(rate_covid * population / 100000)
  ) %>%
  select(time, flu, rsv, covid) %>%
  inner_join(kinsa_raw, by = "time") %>%
  left_join(ili_raw,   by = "time") %>%
  arrange(time)

# Illustrative "as of" date — early Jan 2025 (near peak of 2024-25 flu season).
# Change this to explore a different point in time.
T_date <- as.Date("2025-01-04")
```

## Figure D1: Signal Correlation Matrix

Pearson correlations on log-transformed weekly values. Shows how strongly
each surveillance signal co-moves with the disease-specific RESP-NET counts.

```{r fig-d1-correlation, fig.width=7, fig.height=6}
cor_df <- signal_data %>%
  select(Kinsa = kinsa, ILI = ili, Influenza = flu, RSV = rsv, `COVID-19` = covid) %>%
  filter(complete.cases(.)) %>%
  mutate(across(everything(), log1p))

cor_mat <- cor(cor_df, use = "complete.obs")

cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r") %>%
  mutate(
    Var1 = factor(Var1, levels = colnames(cor_mat)),
    Var2 = factor(Var2, levels = rev(colnames(cor_mat)))
  )

corr_mat <- ggplot(cor_long, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 4.5, fontface = "bold") +
  scale_fill_gradient2(low = "#053061", mid = "white", high = "#67001F",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(
    title = "Signal correlation matrix",
    subtitle = "Pearson r on log(1+x) weekly values -- Michigan, 2019-2026"
  ) +
  theme_paper +
  theme(axis.text = element_text(size = 12),
        legend.position = "right",
        panel.grid = element_blank())

ggsave("figures/fig-d1-correlation.png", corr_mat, width = 7, height = 6, dpi = 300)
corr_mat
```

## Figure D2: Observed Signal Time Series

All five signals plotted over the full data range. Since the signals are
on very different scales (Kinsa is percent ill, ILI and RESP-NET are counts),
each is normalized 0–1 within its own range so co-movement is visible.
The raw scale is labelled in each facet strip.

```{r fig-d2-timeseries, fig.width=11, fig.height=9}
series_long <- signal_data %>%
  select(time,
         Kinsa = kinsa, ILI = ili,
         Influenza = flu, RSV = rsv, `COVID-19` = covid) %>%
  pivot_longer(-time, names_to = "signal", values_to = "value") %>%
  mutate(signal = factor(signal, levels = c("Kinsa", "ILI",
                                            "Influenza", "RSV", "COVID-19")))

# Compute raw range label per signal for the strip
range_labels <- series_long %>%
  group_by(signal) %>%
  summarise(
    lo = floor(min(value, na.rm = TRUE)),
    hi = ceiling(max(value, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(range_label = paste0(signal, "\n[", lo, " – ", hi, "]"),
         signal = factor(signal, levels = levels(series_long$signal)))

series_long <- series_long %>%
  group_by(signal) %>%
  mutate(value_norm = (value - min(value, na.rm = TRUE)) /
                      (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(range_labels %>% select(signal, range_label), by = "signal") %>%
  mutate(range_label = factor(range_label,
                              levels = range_labels$range_label[order(range_labels$signal)]))

time_series_all_signals <- ggplot(series_long, aes(x = time, y = value_norm, color = signal)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  facet_wrap(~ range_label, ncol = 1, scales = "fixed") +
  scale_color_manual(values = signal_colors_all, guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  scale_y_continuous(breaks = c(0, 0.5, 1), labels = c("min", "mid", "max")) +
  labs(
    title = "Observed signal time series — Michigan, full study period",
    subtitle = "Each signal normalized 0–1 | Raw range shown in facet label",
    x = NULL, y = "Normalized value"
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(size = 11, face = "bold"))

ggsave("figures/fig-d2-timeseries.png", time_series_all_signals, width = 11, height = 9, dpi = 300)
time_series_all_signals
```


## Figure D3: Reporting Availability as of `r T_date`

All five signals shown together, each in its own facet. Solid lines =
reliable data as of T. Dashed lines = data that exists but has not yet
matured. Shaded region = the unreliable window for that signal.

```{r fig-d2-delay, fig.width=10, fig.height=11}
window_weeks <- 30
plot_start   <- T_date - weeks(window_weeks)

# Cutoff = last date considered reliable for each signal as of T_date
signal_cutoffs <- c(
  Kinsa      = T_date,              # real-time, no lag
  ILI        = T_date - days(7),   # 1-week reporting lag
  Influenza  = T_date - days(21),  # RESP-NET: ~3-week maturation
  RSV        = T_date - days(21),
  `COVID-19` = T_date - days(21)
)

signal_order <- names(signal_cutoffs)

# Build long data for all 5 signals
all_signals_long <- signal_data %>%
  filter(time >= plot_start, time <= T_date) %>%
  select(time,
         Kinsa = kinsa, ILI = ili,
         Influenza = flu, RSV = rsv, `COVID-19` = covid) %>%
  pivot_longer(-time, names_to = "signal", values_to = "value") %>%
  mutate(signal = factor(signal, levels = signal_order)) %>%
  left_join(
    tibble(signal = factor(signal_order, levels = signal_order),
           cutoff = as.Date(signal_cutoffs)),
    by = "signal"
  ) %>%
  mutate(reliable = time <= cutoff) %>%
  group_by(signal) %>%
  mutate(value_norm = (value - min(value, na.rm = TRUE)) /
                      (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))) %>%
  ungroup()

# Shading rectangles — one per facet
shade_df <- tibble(
  signal     = factor(signal_order, levels = signal_order),
  shade_xmin = as.Date(signal_cutoffs),
  shade_xmax = T_date
)


# Lag annotation labels per facet
lag_labels <- tibble(
  signal = factor(signal_order, levels = signal_order),
  label  = c("No lag (real-time)",
             "~1 week lag",
             "~3 week maturation",
             "~3 week maturation",
             "~3 week maturation")
)

reporting_delay_plot <- ggplot(all_signals_long, aes(x = time, y = value_norm)) +
  geom_rect(data = shade_df,
            aes(xmin = shade_xmin, xmax = shade_xmax,
                ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE,
            fill = "gray85", alpha = 0.7) +
  # Solid = reliable
  geom_line(data = ~ filter(.x, reliable),
            aes(color = signal), linewidth = 1.1) +
  # Dashed = exists but not yet reliable (starts at cutoff to connect)
  geom_line(data = ~ group_by(.x, signal) %>%
              filter(time >= cutoff) %>% ungroup(),
            aes(color = signal), linewidth = 0.9,
            linetype = "dashed", alpha = 0.45) +
  # Vertical line at T
  geom_vline(xintercept = T_date, color = "gray20",
             linetype = "dashed", linewidth = 0.9) +
  # Lag label inside each facet
  geom_text(data = lag_labels,
            aes(label = label),
            x = plot_start + days(3), y = 0.95,
            hjust = 0, size = 3.2, color = "gray30",
            inherit.aes = FALSE) +
  # T label once on the top facet only — avoid repetition
  geom_text(data = tibble(signal = factor("Kinsa", levels = signal_order),
                          time = T_date, value_norm = 1.08),
            aes(x = time, y = value_norm, label = paste("T =", T_date)),
            hjust = 1.05, size = 3.4, color = "gray20", inherit.aes = FALSE) +
  facet_wrap(~ signal, ncol = 1, scales = "free_y") +
  scale_color_manual(values = signal_colors_all, guide = "none") +
  scale_x_date(date_breaks = "4 weeks", date_labels = "%b %d") +
  labs(
    title = "Signal reporting availability as of T",
    subtitle = paste0(
      "Solid = reliable as of T  |  Dashed = exists but not yet mature  |  ",
      "Shaded = unreliable window\n",
      "Kinsa: real-time  \u2502  ILI: ~1 week lag  \u2502  RESP-NET (Flu/RSV/COVID-19): ~3 week maturation"
    ),
    x = NULL, y = "Normalized value (0\u20131 within signal)"
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(size = 12, face = "bold"))

ggsave("figures/fig-d3-reporting-delay.png", reporting_delay_plot,
       width = 10, height = 11, dpi = 300)
reporting_delay_plot
```

---

# Figure 1: Overall CRPS by Model and Disease

```{r fig1-overall-wis, fig.width=10, fig.height=6}
overall <- results %>% #filter(horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(CRPS = mean(crps), .groups = "drop")

fig1_overall <- ggplot(overall, aes(x = reorder(model, CRPS), y = CRPS, fill = model)) +
  geom_col(alpha = 0.85, width = 0.7) +
  facet_wrap(~ disease, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = model_colors, guide = "none") +
  labs(title = "Mean CRPS by Model and Disease",
       subtitle = "100-week rolling-origin backtest | lower is better",
       x = NULL, y = "Mean CRPS") +
  theme_paper

ggsave("figures/fig1-overall-crps.png", fig1_overall,
       width = 10, height = 6, dpi = 300)
fig1_overall
```

# Figure 2: Nowcast Traces (Horizon 0 Only)

One panel per disease showing each model's median prediction vs truth,
restricted to the nowcast horizon (the real-world use case).

```{r fig2-traces, fig.width=11, fig.height=10}
h0 <- results %>% filter(horizon == 0)

# Show the 4 Kinsa models to avoid overplotting
trace_models <- c("Base + Kinsa", "M1 (Diseases \u2192 Kinsa)",
                   "M2 (Kinsa \u2192 Diseases)", "M3 MVN + Kinsa")

p_traces <- h0 %>%
  filter(model %in% trace_models) %>%
  ggplot(aes(x = target_date)) +
  geom_point(aes(y = truth), color = "black", size = 1.5, alpha = 0.6) +
  geom_line(aes(y = median, color = model), linewidth = 0.7) +
  facet_wrap(~ disease, ncol = 1, scales = "free_y") +
  scale_color_manual(values = model_colors) +
  scale_fill_manual(values = model_colors) +
  labs(title = "Nowcast predictions vs. observed counts (horizon 0)",
       subtitle = "Points = truth, lines = posterior median",
       x = NULL, y = "Count") +
  theme_paper +
  guides(fill = "none",
         color = guide_legend(nrow = 2, override.aes = list(linewidth = 1.5)))

ggsave("figures/fig2-nowcast-traces.png", p_traces,
       width = 11, height = 10, dpi = 300)
p_traces
```

# Figure 3: CRPSS — Skill Score vs. Base Model

The CRPSS (Continuous Ranked Probability Skill Score) measures how much
each model improves over the base model: CRPSS = 1 - CRPS_model / CRPS_base.
Values above 0 = improvement; below 0 = worse than base. Being dimensionless,
CRPSS is directly comparable across diseases regardless of their raw scale.
Each dot is one (model, as_of date) pair — the boxplot shows the distribution
of weekly skill scores.

```{r fig3-crpss, fig.width=11, fig.height=7}
# Separate reference baselines: Kinsa models vs base_kinsa, ILI vs base_ili
base_kinsa_crps <- results %>%
  filter(model == "Base + Kinsa", horizon == 0) %>%
  select(disease, target_date, horizon, crps_base = crps)

base_ili_crps <- results %>%
  filter(model == "Base + ILI", horizon == 0) %>%
  select(disease, target_date, horizon, crps_base = crps)

crpss_df <- bind_rows(
  results %>%
    filter(str_detect(model, "Kinsa"),
           !model %in% c("Base + Kinsa")) %>%
    inner_join(base_kinsa_crps, by = c("disease", "target_date", "horizon")),
  results %>%
    filter(str_detect(model, "ILI"),
           !model %in% c("Base + ILI")) %>%
    inner_join(base_ili_crps, by = c("disease", "target_date", "horizon"))
) %>%
  mutate(crpss = 1 - crps / crps_base)

# Summary for ordering
crpss_summary <- crpss_df %>%
  group_by(model, disease) %>%
  summarise(mean_crpss = mean(crpss), .groups = "drop")

fig3_crpss <- ggplot(crpss_df, aes(x = model, y = crpss, fill = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, outlier.size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 2.5, fill = "white", color = "black") +
  facet_wrap(~ disease, scales = "free_y") +
  scale_fill_manual(values = model_colors, guide = "none") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(
    title = "CRPSS relative to family baseline",
    subtitle = paste0(
      "Kinsa models vs. Base + Kinsa  \u2502  ILI models vs. Base + ILI\n",
      "Above 0 = improvement over baseline | Diamond = mean | Box = IQR"
    ),
    x = NULL, y = "CRPSS (higher is better)"
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 37, hjust = 1, size = 10))

ggsave("figures/fig3-crpss.png", fig3_crpss,
       width = 11, height = 7, dpi = 300)
fig3_crpss
```

# Figure 4: CRPS by Horizon

Shows how prediction quality degrades as you move from 2-weeks-back
(easiest) to the nowcast week (hardest).

```{r fig3-horizon, fig.width=10, fig.height=6}
horizon_summary <- results %>%
  group_by(model, disease, horizon) %>%
  summarise(CRPS = mean(crps), .groups = "drop") %>%
  mutate(horizon_label = factor(
    horizon,
    levels = c(-2, -1, 0),
    labels = c("-2\n(2wk back)", "-1\n(1wk back)", "0\n(nowcast)")
  ))

fig4_horizon <- ggplot(horizon_summary,
       aes(x = horizon_label, y = CRPS, color = model, group = model)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  facet_wrap(~ disease, scales = "free_y") +
  scale_color_manual(values = model_colors) +
  labs(title = "Mean CRPS by nowcast horizon",
       subtitle = "Horizon 0 = target week (hardest); -2 = two weeks back (easiest)",
       x = "Horizon", y = "Mean CRPS") +
  theme_paper +
  guides(color = guide_legend(nrow = 2, override.aes = list(linewidth = 1.5)))

ggsave("figures/fig4-horizon.png", fig4_horizon,
       width = 10, height = 6, dpi = 300)
fig4_horizon
```

# Figure 5: Seasonal Phase Performance

How models perform during rising, peak, falling, and trough phases.

```{r fig5-phase, fig.width=15, fig.height=6}
# Classify phases from truth trajectory
phase_labels <- results %>%
  filter(model == levels(results$model)[1]) %>%
  select(disease, target_date, truth) %>%
  distinct() %>%
  group_by(disease) %>%
  arrange(target_date) %>%
  mutate(
    delta_back  = truth - lag(truth, 2),
    delta_fwd   = lead(truth, 2) - truth,
    phase = case_when(
      is.na(delta_back) | is.na(delta_fwd) ~ NA_character_,
      delta_back > 0 & delta_fwd >= 0      ~ "Rising",
      delta_back >= 0 & delta_fwd < 0      ~ "Peak",
      delta_back < 0 & delta_fwd <= 0      ~ "Falling",
      delta_back < 0 & delta_fwd > 0       ~ "Trough",
      TRUE                                  ~ "Stable"
    )
  ) %>%
  ungroup() %>%
  select(disease, target_date, phase) %>%
  filter(!is.na(phase))

phase_crps <- results %>%
  filter(horizon == 0) %>%
  inner_join(phase_labels, by = c("disease", "target_date")) %>%
  group_by(model, disease, phase) %>%
  summarise(
    mean_crps = mean(crps),
    se_crps   = sd(crps) / sqrt(n()),
    n         = n(),
    .groups   = "drop"
  ) %>%
  mutate(phase = factor(phase, levels = c("Rising", "Peak", "Falling", "Trough", "Stable")))

# Show the 4 Kinsa models + M3 ILI (drop base_ili since ≈ base_kinsa)
phase_models <- c("Base + Kinsa",
                   "M1 (Diseases \u2192 Kinsa)", "M2 (Kinsa \u2192 Diseases)",
                   "M3 MVN + Kinsa", "M3 MVN + ILI")

# Shared y-range across all three diseases (for same-scale comparison)
phase_ylim <- phase_crps %>%
  filter(model %in% phase_models) %>%
  summarise(lo = min(mean_crps - se_crps, na.rm = TRUE),
            hi = max(mean_crps + se_crps, na.rm = TRUE))

# Helper: one panel per disease, wider dodge to avoid overlap
make_phase_plot <- function(disease_name) {
  phase_crps %>%
    filter(model %in% phase_models, disease == disease_name) %>%
    ggplot(aes(x = phase, y = mean_crps,
               color = model, shape = model, group = model)) +
    geom_errorbar(aes(ymin = mean_crps - se_crps,
                      ymax = mean_crps + se_crps),
                  position = position_dodge(width = 0.75),
                  width = 0.25, linewidth = 0.65, alpha = 0.85) +
    geom_point(position = position_dodge(width = 0.75),
               size = 3, stroke = 0.9) +
    scale_color_manual(values = model_colors) +
    scale_shape_manual(values = c(16, 17, 15, 18, 8)) +
    coord_cartesian(ylim = c(phase_ylim$lo, phase_ylim$hi)) +
    labs(title = disease_name, x = NULL, y = "Mean CRPS") +
    theme_paper +
    theme(legend.position = "none",
          plot.title = element_text(size = 13, face = "bold"))
}

p_flu   <- make_phase_plot("Influenza")
p_rsv   <- make_phase_plot("RSV")
p_covid <- make_phase_plot("COVID-19")

fig5_phase <- (p_flu | p_rsv | p_covid) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title    = "Mean CRPS by seasonal phase (horizon 0)",
    subtitle = "Points = mean CRPS \u2502 Error bars = \u00b11 SE \u2502 Phase classified from 2-week forward/backward slope",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray30", size = 11)
    )
  ) &
  theme(legend.position = "bottom",
        legend.title     = element_blank())

ggsave("figures/fig5-phase.png", fig5_phase,
       width = 15, height = 6, dpi = 300)
fig5_phase
```

# Figure 6: Performance by Respiratory Season

Groups the 100 backtest weeks by respiratory season (Oct–Sep), so we can
see whether model rankings are stable across years or whether particular
models shine in certain seasons (e.g., the 2024–25 flu peak).

```{r fig-season, fig.width=13, fig.height=7}
results_season <- results %>%
  filter(horizon == 0) %>%
  mutate(
    # Respiratory season: Oct 2024 - Sep 2025 = "2024-25"
    resp_year = if_else(month(target_date) >= 10,
                        year(target_date),
                        year(target_date) - 1),
    season = paste0(resp_year, "-",
                    str_sub(as.character(resp_year + 1), 3, 4))
  )

season_metrics <- results_season %>%
  group_by(model, disease, season) %>%
  summarise(
    n         = n(),
    mean_crps = mean(crps),
    se_crps   = sd(crps) / sqrt(n()),
    .groups   = "drop"
  )

# Drop seasons with too few weeks in the backtest window (e.g. partial edges)
season_metrics <- season_metrics %>%
  group_by(season) %>% filter(sum(n) >= 5) %>% ungroup()

fig6_season <- ggplot(season_metrics,
       aes(x = season, y = mean_crps,
           color = model, shape = model, group = model)) +
  geom_errorbar(aes(ymin = mean_crps - se_crps,
                    ymax = mean_crps + se_crps),
                position = position_dodge(width = 0.7),
                width = 0.25, linewidth = 0.6, alpha = 0.85) +
  geom_point(position = position_dodge(width = 0.7),
             size = 3, stroke = 0.9) +
  facet_wrap(~ disease, scales = "free_y") +
  scale_color_manual(values = model_colors) +
  scale_shape_manual(values = c(16, 17, 15, 18, 1, 2, 0, 5)) +
  labs(
    title = "Mean CRPS by model, disease, and respiratory season (horizon 0)",
    subtitle = "Respiratory season = Oct\u2013Sep  \u2502  Points = mean CRPS  \u2502  Error bars = \u00b11 SE",
    x = "Respiratory season", y = "Mean CRPS"
  ) +
  theme_paper +
  guides(color = guide_legend(nrow = 2, override.aes = list(size = 3)),
         shape = guide_legend(nrow = 2)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggsave("figures/fig6-season.png", fig6_season,
       width = 13, height = 7, dpi = 300)
fig6_season
```

## Table: CRPS by Respiratory Season

```{r table-season}
season_tbl <- season_metrics %>%
  mutate(CRPS = round(mean_crps, 1)) %>%
  select(season, model, disease, CRPS) %>%
  pivot_wider(names_from = disease, values_from = CRPS) %>%
  arrange(season, model) %>%
  select(season, model, Influenza, RSV, `COVID-19`)

tbl_season_kbl <- kbl(season_tbl %>% select(-season),
    align = c("l", rep("r", 3)),
    caption = "Mean CRPS by model and respiratory season (horizon 0). Each season covers Oct-Sep.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  pack_rows(index = table(season_tbl$season))

save_kable(tbl_season_kbl, file = "figures/table-season.html", self_contained = TRUE)
tbl_season_kbl
```

# Figure 7: Early-Season Performance (Sep–Dec)

Restricts the backtest to the early-season weeks of each respiratory season
(Sep–Dec). This is arguably the most operationally important window: cases
are climbing, RESP-NET has not yet matured, and public-health decisions
about capacity and messaging depend on getting an early read on the wave.

```{r fig-early-season, fig.width=12, fig.height=6}
early_season <- results %>%
  filter(horizon == 0, month(target_date) %in% 9:12)

early_summary <- early_season %>%
  group_by(model, disease) %>%
  summarise(
    n         = n(),
    mean_crps = mean(crps),
    se_crps   = sd(crps) / sqrt(n()),
    .groups   = "drop"
  )

# Compare directly against the full-year average to highlight
# how model rankings shift in the ramp-up phase.
full_year_summary <- results %>%
  filter(horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(mean_crps = mean(crps), .groups = "drop") %>%
  mutate(window = "Full year")

early_summary_plot <- early_summary %>%
  mutate(window = "Early season (Sep\u2013Dec)") %>%
  select(model, disease, mean_crps, window) %>%
  bind_rows(full_year_summary) %>%
  mutate(window = factor(window,
                         levels = c("Full year", "Early season (Sep\u2013Dec)")))

fig7_early_season <- ggplot(early_summary_plot,
       aes(x = reorder(model, mean_crps), y = mean_crps,
           fill = model, alpha = window)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ disease, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = model_colors, guide = "none") +
  scale_alpha_manual(values = c("Full year" = 0.4,
                                "Early season (Sep\u2013Dec)" = 1),
                     name = NULL) +
  labs(
    title = "Mean CRPS: early season vs. full year (horizon 0)",
    subtitle = "Early-season weeks are when nowcasts matter most \u2014 cases climbing, RESP-NET not yet mature",
    x = NULL, y = "Mean CRPS"
  ) +
  theme_paper

ggsave("figures/fig7-early-season.png", fig7_early_season,
       width = 12, height = 6, dpi = 300)
fig7_early_season
```

## Table: Early-Season Nowcast Performance

```{r table-early-season}
early_tbl <- early_summary %>%
  mutate(CRPS = round(mean_crps, 1),
         SE   = round(se_crps, 2)) %>%
  select(model, disease, n, CRPS, SE) %>%
  pivot_wider(names_from = disease,
              values_from = c(CRPS, SE),
              names_glue = "{disease}_{.value}") %>%
  select(model,
         starts_with("Influenza"),
         starts_with("RSV"),
         starts_with("COVID"))

# Bold best CRPS per disease column
early_fmt <- early_tbl
for (col in names(early_fmt)[-1]) {
  if (grepl("_CRPS$", col)) {
    x <- early_fmt[[col]]
    best <- which.min(x)
    out <- as.character(x)
    out[best] <- cell_spec(out[best], bold = TRUE)
    early_fmt[[col]] <- out
  } else {
    early_fmt[[col]] <- as.character(early_fmt[[col]])
  }
}

tbl_early_kbl <- kbl(early_fmt, escape = FALSE, align = c("l", rep("r", 6)),
    caption = paste(
      "Mean CRPS and standard error during early-season weeks (Sep-Dec, horizon 0).",
      "Bold = best per disease. Standard error computed across weeks."
    )) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  add_header_above(c(" " = 1, "Influenza" = 2, "RSV" = 2, "COVID-19" = 2))

save_kable(tbl_early_kbl, file = "figures/table-early-season.html", self_contained = TRUE)
tbl_early_kbl
```

---

# Tables

```{r table-helpers}
# Base CRPS for skill score computation (family-matched)
base_crps_kinsa <- results %>%
  filter(model == "Base + Kinsa") %>%
  select(disease, target_date, horizon, crps_base = crps)

base_crps_ili <- results %>%
  filter(model == "Base + ILI") %>%
  select(disease, target_date, horizon, crps_base = crps)

results_with_skill <- bind_rows(
  results %>%
    filter(str_detect(model, "Kinsa")) %>%
    left_join(base_crps_kinsa, by = c("disease", "target_date", "horizon")),
  results %>%
    filter(str_detect(model, "ILI")) %>%
    left_join(base_crps_ili, by = c("disease", "target_date", "horizon"))
) %>%
  mutate(crpss = 1 - crps / crps_base)

# Table 1 restricted to horizon 0 (nowcast week) — the primary use case
metrics_all <- results_with_skill %>%
  filter(horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(
    n           = n(),
    mean_crps   = mean(crps),
    mean_crpss  = mean(crpss),
    cover_95    = mean(truth >= q025 & truth <= q975) * 100,
    .groups     = "drop"
  )
```

## Table 1: Overall Nowcast Accuracy

```{r table-1}
tbl1 <- metrics_all %>%
  mutate(
    CRPS      = round(mean_crps,  1),
    CRPSS     = round(mean_crpss, 3),
    `95% Cov` = round(cover_95,   0)
  ) %>%
  select(model, disease, CRPS, CRPSS, `95% Cov`)

tbl1_wide <- tbl1 %>%
  pivot_wider(names_from = disease,
              values_from = c(CRPS, CRPSS, `95% Cov`),
              names_glue = "{disease}_{.value}") %>%
  select(model, starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

# Bold best per column
bold_best <- function(x, target = NA) {
  best <- if (is.na(target)) which.min(x) else which.min(abs(x - target))
  out <- as.character(x)
  out[best] <- cell_spec(out[best], bold = TRUE)
  out
}

# Apply bolding column-by-column to avoid earlier columns becoming character
# before later columns are processed
tbl1_fmt <- tbl1_wide
for (col in names(tbl1_fmt)[-1]) {
  if (grepl("Cov", col)) {
    tbl1_fmt[[col]] <- bold_best(tbl1_fmt[[col]], target = 95)
  } else if (grepl("CRPSS", col)) {
    # For CRPSS, higher is better — bold the maximum
    x <- tbl1_fmt[[col]]
    best <- which.max(x)
    out <- as.character(x)
    out[best] <- cell_spec(out[best], bold = TRUE)
    tbl1_fmt[[col]] <- out
  } else {
    tbl1_fmt[[col]] <- bold_best(tbl1_fmt[[col]])  # lower is better
  }
}

tbl1_kbl <- kbl(tbl1_fmt, escape = FALSE, align = c("l", rep("r", 9)),
    caption = "Table 1: Nowcast performance at horizon 0 (100-week rolling-origin backtest). CRPS is sample-based from full posterior draws. CRPSS = 1 - CRPS/CRPS\\_base (vs. family baseline; higher is better). Bold = best per column.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  add_header_above(c(" " = 1, "Influenza" = 3, "RSV" = 3, "COVID-19" = 3))

save_kable(tbl1_kbl, file = "figures/table1-overall.html", self_contained = TRUE)
tbl1_kbl
```

## Table 2: CRPSS by Model and Disease

CRPSS = 1 - CRPS_model / CRPS_base. Kinsa models compared against
Base + Kinsa; ILI models compared against Base + ILI. "% Weeks Better"
= percentage of the 100 backtest weeks where this model beat its baseline.

```{r table-2}
skill_tbl <- results_with_skill %>%
  filter(!model %in% c("Base + Kinsa", "Base + ILI"), horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(
    CRPSS      = round(mean(crpss), 3),
    `% Better` = round(mean(crpss > 0) * 100, 0),
    .groups    = "drop"
  )

skill_wide <- skill_tbl %>%
  pivot_wider(names_from = disease,
              values_from = c(CRPSS, `% Better`),
              names_glue = "{disease}_{.value}") %>%
  select(model, starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

# Bold highest CRPSS per disease column
skill_fmt <- skill_wide
for (col in names(skill_fmt)[-1]) {
  x <- skill_fmt[[col]]
  if (grepl("CRPSS", col)) {
    best <- which.max(x)
  } else {
    best <- which.max(x)   # % Better: also higher is better
  }
  out <- as.character(x)
  out[best] <- cell_spec(out[best], bold = TRUE)
  skill_fmt[[col]] <- out
}

tbl2_kbl <- kbl(skill_fmt, escape = FALSE, align = c("l", rep("r", 6)),
    caption = "Table 2: CRPSS relative to family baseline (Kinsa models vs. Base + Kinsa; ILI models vs. Base + ILI). Higher CRPSS = greater improvement. Bold = best per column. Horizon = 0") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  add_header_above(c(" " = 1, "Influenza" = 2, "RSV" = 2, "COVID-19" = 2))

save_kable(tbl2_kbl, file = "figures/table2-crpss.html", self_contained = TRUE)
tbl2_kbl
```

## Table 3: Performance by Horizon

```{r table-3}
horizon_tbl <- results %>%
  group_by(model, disease, horizon) %>%
  summarise(CRPS = round(mean(crps), 1), .groups = "drop") %>%
  mutate(horizon_label = factor(
    horizon, levels = c(-2, -1, 0),
    labels = c("Horizon -2 (2wk back)", "Horizon -1 (1wk back)", "Horizon 0 (nowcast)")
  ))

horizon_wide <- horizon_tbl %>%
  pivot_wider(names_from = disease, values_from = CRPS) %>%
  arrange(horizon, model) %>%
  select(horizon_label, model, Influenza, RSV, `COVID-19`)

tbl3_kbl <- kbl(horizon_wide %>% select(-horizon_label),
    align = c("l", rep("r", 3)),
    caption = "Table 3: Mean CRPS by model and horizon. Horizon 0 is the nowcast week (hardest).") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  pack_rows(index = table(horizon_wide$horizon_label))

save_kable(tbl3_kbl, file = "figures/table3-horizon.html", self_contained = TRUE)
tbl3_kbl
```

## Table 4: CRPS at Horizon 0 (Nowcast Week)

Sample-based CRPS at horizon 0 only — the hardest and most relevant
prediction task. Kinsa and ILI model families shown side by side.

```{r table-4-crps}
crps_tbl <- results %>%
  filter(horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(CRPS = round(mean(crps), 1), .groups = "drop") %>%
  pivot_wider(names_from = disease, values_from = CRPS) %>%
  mutate(
    family    = if_else(str_detect(model, "Kinsa"), "Kinsa", "ILI"),
    structure = case_when(
      str_detect(model, "Base") ~ "Base",
      str_detect(model, "M1")   ~ "M1",
      str_detect(model, "M2")   ~ "M2",
      str_detect(model, "M3")   ~ "M3"
    )
  ) %>%
  arrange(structure, family) %>%
  select(structure, family, model, Influenza, RSV, `COVID-19`)

# Bold lowest CRPS per (disease, structure) group
crps_fmt <- crps_tbl
for (col in c("Influenza", "RSV", "COVID-19")) {
  vals <- crps_fmt[[col]]
  for (struct in unique(crps_fmt$structure)) {
    idx <- which(crps_fmt$structure == struct)
    best <- idx[which.min(vals[idx])]
    crps_fmt[[col]][best] <- cell_spec(as.character(vals[best]), bold = TRUE)
    crps_fmt[[col]][setdiff(idx, best)] <- as.character(vals[setdiff(idx, best)])
  }
}

tbl4_kbl <- kbl(crps_fmt %>% select(-structure),
    escape = FALSE, align = c("l", "l", "r", "r", "r"),
    caption = paste(
      "Table 4: Mean CRPS at horizon 0 (nowcast week) by model and disease.",
      "CRPS is sample-based from full posterior draws (~6,000 samples per fit).",
      "Bold = best within each model structure (Kinsa vs. ILI)."
    )) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  pack_rows(index = table(crps_fmt$structure)) %>%
  add_header_above(c(" " = 2, "Mean CRPS" = 3))

save_kable(tbl4_kbl, file = "figures/table4-horizon0-crps.html", self_contained = TRUE)
tbl4_kbl
```
