---
title: "Paper Figures — Bayesian Respiratory Disease Nowcasting"
output:
  html_document:
    fig_width: 10
    fig_height: 7
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.retina = 2, dpi = 300
)
```

```{r load}
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)

# ---- Load cached backtest results ----
results <- read_csv("results/backtest_predictions.csv", show_col_types = FALSE) %>%
  mutate(as_of = as.Date(as_of), target_date = as.Date(target_date))

# ---- Labels and colors ----
model_labels <- c(
  base_kinsa = "Base + Kinsa",
  m1_kinsa   = "M1 (Diseases \u2192 Kinsa)",
  m2_kinsa   = "M2 (Kinsa \u2192 Diseases)",
  m3_kinsa   = "M3 MVN + Kinsa",
  base_ili   = "Base + ILI",
  m1_ili     = "M1 (Diseases \u2192 ILI)",
  m2_ili     = "M2 (ILI \u2192 Diseases)",
  m3_ili     = "M3 MVN + ILI"
)

disease_labels <- c(flu = "Influenza", rsv = "RSV", covid = "COVID-19")

# 8 distinguishable colors: Kinsa family = blues, ILI family = oranges/reds
model_colors <- c(
  "Base + Kinsa"             = "#4393C3",
  "M1 (Diseases \u2192 Kinsa)"  = "#2166AC",
  "M2 (Kinsa \u2192 Diseases)"  = "#053061",
  "M3 MVN + Kinsa"           = "#92C5DE",
  "Base + ILI"               = "#F4A582",
  "M1 (Diseases \u2192 ILI)"    = "#D6604D",
  "M2 (ILI \u2192 Diseases)"    = "#B2182B",
  "M3 MVN + ILI"             = "#FDDBC7"
)

results <- results %>%
  mutate(
    model   = recode(model, !!!model_labels),
    disease = recode(disease, !!!disease_labels),
    model   = factor(model, levels = model_labels)
  )

# Common theme
theme_paper <- theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title    = element_blank(),
    strip.text      = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "gray30", size = 11)
  )
```

# Descriptive Signal Analysis

```{r raw-data-load, message=FALSE, warning=FALSE}
library(tidycensus)

kinsa_path  <- "/Users/tobysalmon/Desktop/kinsa_michigan_weekly.csv"
fips        <- "26"
kinsa_state <- "Michigan"

resp_raw <- read_csv(
  "https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz",
  show_col_types = FALSE
) %>% filter(geography == fips, age == "Total")

ili_raw <- read_csv(
  "https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/delphi_ili_fluview/standard/data.csv.gz",
  show_col_types = FALSE
) %>%
  filter(geography == fips) %>%
  select(time, ili = delphi_fluview_num_ili)

kinsa_raw <- read_csv(kinsa_path, show_col_types = FALSE) %>%
  filter(region_name == kinsa_state) %>%
  mutate(time = as.Date(time)) %>%
  select(time, kinsa = percent_ill)

# Population (same vintage as Backtest.Rmd)
state_pre <- get_estimates(geography = "state", state = fips,
  product = "population", vintage = 2019, time_series = TRUE) |>
  filter(variable == "POP") |> mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |> arrange(year) |> pull(value)
state_pop <- get_estimates(geography = "state", state = fips,
  product = "population", vintage = 2025, time_series = TRUE) |>
  filter(variable == "POPESTIMATE", year %in% 2020:2025) |>
  arrange(year) |> pull(value)
pop_tbl <- tibble(year = 2018:2026,
                  population = c(state_pre, state_pop, state_pop[6]))

signal_data <- resp_raw %>%
  mutate(year = year(time)) %>%
  left_join(pop_tbl, by = "year") %>%
  mutate(
    flu   = round(rate_flu   * population / 100000),
    rsv   = round(rate_rsv   * population / 100000),
    covid = round(rate_covid * population / 100000)
  ) %>%
  select(time, flu, rsv, covid) %>%
  inner_join(kinsa_raw, by = "time") %>%
  left_join(ili_raw,   by = "time") %>%
  arrange(time)

# Illustrative "as of" date — early Jan 2025 (near peak of 2024-25 flu season).
# Change this to explore a different point in time.
T_date <- as.Date("2025-01-04")
```

## Figure D1: Signal Correlation Matrix

Pearson correlations on log-transformed weekly values. Shows how strongly
each surveillance signal co-moves with the disease-specific RESP-NET counts.

```{r fig-d1-correlation, fig.width=7, fig.height=6}
cor_df <- signal_data %>%
  select(Kinsa = kinsa, ILI = ili, Influenza = flu, RSV = rsv, `COVID-19` = covid) %>%
  filter(complete.cases(.)) %>%
  mutate(across(everything(), log1p))

cor_mat <- cor(cor_df, use = "complete.obs")

cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r") %>%
  mutate(
    Var1 = factor(Var1, levels = colnames(cor_mat)),
    Var2 = factor(Var2, levels = rev(colnames(cor_mat)))
  )

ggplot(cor_long, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 4.5, fontface = "bold") +
  scale_fill_gradient2(low = "#053061", mid = "white", high = "#67001F",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(
    title = "Signal correlation matrix",
    subtitle = "Pearson r on log\u2081\u208a(x) weekly values \u2014 Michigan, 2019\u20132026"
  ) +
  theme_paper +
  theme(axis.text = element_text(size = 12),
        legend.position = "right",
        panel.grid = element_blank())
```

## Figure D2: Observed Signal Time Series

All five signals plotted over the full data range. Since the signals are
on very different scales (Kinsa is percent ill, ILI and RESP-NET are counts),
each is normalized 0–1 within its own range so co-movement is visible.
The raw scale is labelled in each facet strip.

```{r fig-d2-timeseries, fig.width=11, fig.height=9}
series_long <- signal_data %>%
  select(time,
         Kinsa = kinsa, ILI = ili,
         Influenza = flu, RSV = rsv, `COVID-19` = covid) %>%
  pivot_longer(-time, names_to = "signal", values_to = "value") %>%
  mutate(signal = factor(signal, levels = c("Kinsa", "ILI",
                                            "Influenza", "RSV", "COVID-19")))

# Compute raw range label per signal for the strip
range_labels <- series_long %>%
  group_by(signal) %>%
  summarise(
    lo = floor(min(value, na.rm = TRUE)),
    hi = ceiling(max(value, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(range_label = paste0(signal, "\n[", lo, " – ", hi, "]"),
         signal = factor(signal, levels = levels(series_long$signal)))

series_long <- series_long %>%
  group_by(signal) %>%
  mutate(value_norm = (value - min(value, na.rm = TRUE)) /
                      (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(range_labels %>% select(signal, range_label), by = "signal") %>%
  mutate(range_label = factor(range_label,
                              levels = range_labels$range_label[order(range_labels$signal)]))

ggplot(series_long, aes(x = time, y = value_norm, color = signal)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  facet_wrap(~ range_label, ncol = 1, scales = "fixed") +
  scale_color_manual(values = signal_colors_all, guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  scale_y_continuous(breaks = c(0, 0.5, 1), labels = c("min", "mid", "max")) +
  labs(
    title = "Observed signal time series — Michigan, full study period",
    subtitle = "Each signal normalized 0–1 | Raw range shown in facet label",
    x = NULL, y = "Normalized value"
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(size = 11, face = "bold"))
```

## Figure D3: Reporting Availability as of `r T_date`

All five signals shown together, each in its own facet. Solid lines =
reliable data as of T. Dashed lines = data that exists but has not yet
matured. Shaded region = the unreliable window for that signal.

```{r fig-d2-delay, fig.width=10, fig.height=11}
window_weeks <- 30
plot_start   <- T_date - weeks(window_weeks)

# Cutoff = last date considered reliable for each signal as of T_date
signal_cutoffs <- c(
  Kinsa      = T_date,              # real-time, no lag
  ILI        = T_date - days(7),   # 1-week reporting lag
  Influenza  = T_date - days(21),  # RESP-NET: ~3-week maturation
  RSV        = T_date - days(21),
  `COVID-19` = T_date - days(21)
)

signal_order <- names(signal_cutoffs)

# Build long data for all 5 signals
all_signals_long <- signal_data %>%
  filter(time >= plot_start, time <= T_date) %>%
  select(time,
         Kinsa = kinsa, ILI = ili,
         Influenza = flu, RSV = rsv, `COVID-19` = covid) %>%
  pivot_longer(-time, names_to = "signal", values_to = "value") %>%
  mutate(signal = factor(signal, levels = signal_order)) %>%
  left_join(
    tibble(signal = factor(signal_order, levels = signal_order),
           cutoff = as.Date(signal_cutoffs)),
    by = "signal"
  ) %>%
  mutate(reliable = time <= cutoff) %>%
  group_by(signal) %>%
  mutate(value_norm = (value - min(value, na.rm = TRUE)) /
                      (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))) %>%
  ungroup()

# Shading rectangles — one per facet
shade_df <- tibble(
  signal     = factor(signal_order, levels = signal_order),
  shade_xmin = as.Date(signal_cutoffs),
  shade_xmax = T_date
)

signal_colors_all <- c(
  Kinsa      = "#E69F00",   # amber/orange
  ILI        = "#CC79A7",   # pink/magenta
  Influenza  = "#0072B2",   # deep blue
  RSV        = "#009E73",   # teal green
  `COVID-19` = "#D55E00"    # vermillion red
)

# Lag annotation labels per facet
lag_labels <- tibble(
  signal = factor(signal_order, levels = signal_order),
  label  = c("No lag (real-time)",
             "~1 week lag",
             "~3 week maturation",
             "~3 week maturation",
             "~3 week maturation")
)

ggplot(all_signals_long, aes(x = time, y = value_norm)) +
  geom_rect(data = shade_df,
            aes(xmin = shade_xmin, xmax = shade_xmax,
                ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE,
            fill = "gray85", alpha = 0.7) +
  # Solid = reliable
  geom_line(data = ~ filter(.x, reliable),
            aes(color = signal), linewidth = 1.1) +
  # Dashed = exists but not yet reliable (starts at cutoff to connect)
  geom_line(data = ~ group_by(.x, signal) %>%
              filter(time >= cutoff) %>% ungroup(),
            aes(color = signal), linewidth = 0.9,
            linetype = "dashed", alpha = 0.45) +
  # Vertical line at T
  geom_vline(xintercept = T_date, color = "gray20",
             linetype = "dashed", linewidth = 0.9) +
  # Lag label inside each facet
  geom_text(data = lag_labels,
            aes(label = label),
            x = plot_start + days(3), y = 0.95,
            hjust = 0, size = 3.2, color = "gray30",
            inherit.aes = FALSE) +
  # T label once on the top facet only — avoid repetition
  geom_text(data = tibble(signal = factor("Kinsa", levels = signal_order),
                          time = T_date, value_norm = 1.08),
            aes(x = time, y = value_norm, label = paste("T =", T_date)),
            hjust = 1.05, size = 3.4, color = "gray20", inherit.aes = FALSE) +
  facet_wrap(~ signal, ncol = 1, scales = "free_y") +
  scale_color_manual(values = signal_colors_all, guide = "none") +
  scale_x_date(date_breaks = "4 weeks", date_labels = "%b %d") +
  labs(
    title = "Signal reporting availability as of T",
    subtitle = paste0(
      "Solid = reliable as of T  |  Dashed = exists but not yet mature  |  ",
      "Shaded = unreliable window\n",
      "Kinsa: real-time  \u2502  ILI: ~1 week lag  \u2502  RESP-NET (Flu/RSV/COVID-19): ~3 week maturation"
    ),
    x = NULL, y = "Normalized value (0\u20131 within signal)"
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(size = 12, face = "bold"))
```

---

# Figure 1: Overall CRPS by Model and Disease

```{r fig1-overall-wis, fig.width=10, fig.height=6}
overall <- results %>%
  group_by(model, disease) %>%
  summarise(CRPS = mean(crps), .groups = "drop")

ggplot(overall, aes(x = reorder(model, CRPS), y = CRPS, fill = model)) +
  geom_col(alpha = 0.85, width = 0.7) +
  facet_wrap(~ disease, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = model_colors, guide = "none") +
  labs(title = "Mean CRPS by model and disease",
       subtitle = "100-week rolling-origin backtest | lower is better",
       x = NULL, y = "Mean CRPS") +
  theme_paper
```

# Figure 2: Nowcast Traces (Horizon 0 Only)

One panel per disease showing each model's median prediction vs truth,
restricted to the nowcast horizon (the real-world use case).

```{r fig2-traces, fig.width=11, fig.height=10}
h0 <- results %>% filter(horizon == 0)

# Show the 4 Kinsa models to avoid overplotting
trace_models <- c("Base + Kinsa", "M1 (Diseases \u2192 Kinsa)",
                   "M2 (Kinsa \u2192 Diseases)", "M3 MVN + Kinsa")

p_traces <- h0 %>%
  filter(model %in% trace_models) %>%
  ggplot(aes(x = target_date)) +
  geom_point(aes(y = truth), color = "black", size = 1.5, alpha = 0.6) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = model), alpha = 0.15) +
  geom_line(aes(y = median, color = model), linewidth = 0.7) +
  facet_wrap(~ disease, ncol = 1, scales = "free_y") +
  scale_color_manual(values = model_colors) +
  scale_fill_manual(values = model_colors) +
  labs(title = "Nowcast predictions vs. observed counts (horizon 0)",
       subtitle = "Points = truth, lines = posterior median, ribbon = 50% CI",
       x = NULL, y = "Count") +
  theme_paper +
  guides(fill = "none",
         color = guide_legend(nrow = 2, override.aes = list(linewidth = 1.5)))

p_traces
```

# Figure 3: CRPSS — Skill Score vs. Base Model

The CRPSS (Continuous Ranked Probability Skill Score) measures how much
each model improves over the base model: CRPSS = 1 - CRPS_model / CRPS_base.
Values above 0 = improvement; below 0 = worse than base. Being dimensionless,
CRPSS is directly comparable across diseases regardless of their raw scale.
Each dot is one (model, as_of date) pair — the boxplot shows the distribution
of weekly skill scores.

```{r fig3-crpss, fig.width=11, fig.height=7}
# Separate reference baselines: Kinsa models vs base_kinsa, ILI vs base_ili
base_kinsa_crps <- results %>%
  filter(model == "Base + Kinsa") %>%
  select(disease, target_date, horizon, crps_base = crps)

base_ili_crps <- results %>%
  filter(model == "Base + ILI") %>%
  select(disease, target_date, horizon, crps_base = crps)

crpss_df <- bind_rows(
  results %>%
    filter(str_detect(model, "Kinsa"),
           !model %in% c("Base + Kinsa")) %>%
    inner_join(base_kinsa_crps, by = c("disease", "target_date", "horizon")),
  results %>%
    filter(str_detect(model, "ILI"),
           !model %in% c("Base + ILI")) %>%
    inner_join(base_ili_crps, by = c("disease", "target_date", "horizon"))
) %>%
  mutate(crpss = 1 - crps / crps_base)

# Summary for ordering
crpss_summary <- crpss_df %>%
  group_by(model, disease) %>%
  summarise(mean_crpss = mean(crpss), .groups = "drop")

ggplot(crpss_df, aes(x = model, y = crpss, fill = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, outlier.size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 2.5, fill = "white", color = "black") +
  facet_wrap(~ disease, scales = "free_y") +
  scale_fill_manual(values = model_colors, guide = "none") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  labs(
    title = "CRPSS relative to family baseline",
    subtitle = paste0(
      "Kinsa models vs. Base + Kinsa  \u2502  ILI models vs. Base + ILI\n",
      "Above 0 = improvement over baseline | Diamond = mean | Box = IQR"
    ),
    x = NULL, y = "CRPSS (higher is better)"
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 10))
```

# Figure 4: CRPS by Horizon

Shows how prediction quality degrades as you move from 2-weeks-back
(easiest) to the nowcast week (hardest).

```{r fig3-horizon, fig.width=10, fig.height=6}
horizon_summary <- results %>%
  group_by(model, disease, horizon) %>%
  summarise(CRPS = mean(crps), .groups = "drop") %>%
  mutate(horizon_label = factor(
    horizon,
    levels = c(-2, -1, 0),
    labels = c("-2\n(2wk back)", "-1\n(1wk back)", "0\n(nowcast)")
  ))

ggplot(horizon_summary,
       aes(x = horizon_label, y = CRPS, color = model, group = model)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  facet_wrap(~ disease, scales = "free_y") +
  scale_color_manual(values = model_colors) +
  labs(title = "Mean CRPS by nowcast horizon",
       subtitle = "Horizon 0 = target week (hardest); -2 = two weeks back (easiest)",
       x = "Horizon", y = "Mean CRPS") +
  theme_paper +
  guides(color = guide_legend(nrow = 2, override.aes = list(linewidth = 1.5)))
```

# Figure 5: Kinsa vs ILI Signal Comparison

Direct comparison of the two signal families. For each model structure
(Base, M1, M2, M3), plot the Kinsa version's WIS against the ILI version's.

```{r fig4-kinsa-vs-ili, fig.width=9, fig.height=6}
# Map each model to its structure and signal family
signal_comparison <- results %>%
  mutate(
    structure = case_when(
      str_detect(model, "Base")  ~ "Base",
      str_detect(model, "M1")    ~ "M1",
      str_detect(model, "M2")    ~ "M2",
      str_detect(model, "M3")    ~ "M3"
    ),
    signal = if_else(str_detect(model, "Kinsa"), "Kinsa", "ILI")
  ) %>%
  group_by(structure, signal, disease) %>%
  summarise(CRPS = mean(crps), .groups = "drop") %>%
  pivot_wider(names_from = signal, values_from = CRPS)

ggplot(signal_comparison, aes(x = ILI, y = Kinsa, shape = structure, color = structure)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 4, stroke = 1.2) +
  facet_wrap(~ disease, scales = "free") +
  scale_shape_manual(values = c(Base = 16, M1 = 17, M2 = 15, M3 = 18)) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Kinsa vs. ILI signal: Mean CRPS comparison",
       subtitle = "Points below the diagonal = Kinsa outperforms ILI",
       x = "Mean CRPS (ILI version)", y = "Mean CRPS (Kinsa version)",
       shape = "Model structure", color = "Model structure") +
  theme_paper +
  theme(legend.position = "right")
```

# Figure 6: Seasonal Phase Performance

How models perform during rising, peak, falling, and trough phases.

```{r fig5-phase, fig.width=10, fig.height=7}
# Classify phases from truth trajectory
phase_labels <- results %>%
  filter(model == levels(results$model)[1]) %>%
  select(disease, target_date, truth) %>%
  distinct() %>%
  group_by(disease) %>%
  arrange(target_date) %>%
  mutate(
    delta_back  = truth - lag(truth, 2),
    delta_fwd   = lead(truth, 2) - truth,
    phase = case_when(
      is.na(delta_back) | is.na(delta_fwd) ~ NA_character_,
      delta_back > 0 & delta_fwd >= 0      ~ "Rising",
      delta_back >= 0 & delta_fwd < 0      ~ "Peak",
      delta_back < 0 & delta_fwd <= 0      ~ "Falling",
      delta_back < 0 & delta_fwd > 0       ~ "Trough",
      TRUE                                  ~ "Stable"
    )
  ) %>%
  ungroup() %>%
  select(disease, target_date, phase) %>%
  filter(!is.na(phase))

phase_crps <- results %>%
  inner_join(phase_labels, by = c("disease", "target_date")) %>%
  group_by(model, disease, phase) %>%
  summarise(CRPS = mean(crps), n = n(), .groups = "drop") %>%
  mutate(phase = factor(phase, levels = c("Rising", "Peak", "Falling", "Trough", "Stable")))

# Show the 4 Kinsa models + M3 ILI (drop base_ili since ≈ base_kinsa)
phase_models <- c("Base + Kinsa",
                   "M1 (Diseases \u2192 Kinsa)", "M2 (Kinsa \u2192 Diseases)",
                   "M3 MVN + Kinsa", "M3 MVN + ILI")

ggplot(phase_crps %>% filter(model %in% phase_models),
       aes(x = phase, y = CRPS, fill = model)) +
  geom_col(position = position_dodge(width = 0.8), alpha = 0.85, width = 0.7) +
  facet_wrap(~ disease, scales = "free_y") +
  scale_fill_manual(values = model_colors) +
  labs(title = "Mean CRPS by seasonal phase",
       subtitle = "Rising = counts increasing, Peak = local max, Falling = decreasing, Trough = local min",
       x = NULL, y = "Mean CRPS") +
  theme_paper +
  guides(fill = guide_legend(nrow = 2))
```

# Figure 7: Calibration

Are prediction intervals well-calibrated? Bars show empirical coverage
vs. nominal levels (dashed lines).

```{r fig6-calibration, fig.width=10, fig.height=5}
cal_models <- c("Base + Kinsa", "M1 (Diseases \u2192 Kinsa)",
                "M2 (Kinsa \u2192 Diseases)", "M3 MVN + Kinsa", "M3 MVN + ILI")

calibration <- results %>%
  filter(model %in% cal_models) %>%
  group_by(model, disease) %>%
  summarise(
    `50%` = mean(truth >= q25  & truth <= q75)  * 100,
    `95%` = mean(truth >= q025 & truth <= q975) * 100,
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`50%`, `95%`),
               names_to = "interval", values_to = "coverage") %>%
  mutate(nominal = as.numeric(str_extract(interval, "\\d+")))

ggplot(calibration, aes(x = model, y = coverage, fill = model)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(aes(yintercept = nominal), linetype = "dashed", color = "gray30") +
  facet_grid(interval ~ disease) +
  scale_fill_manual(values = model_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Empirical coverage vs. nominal prediction intervals",
       subtitle = "Dashed line = target | bars above = conservative, below = overconfident",
       x = NULL, y = "Coverage (%)") +
  theme_paper +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 10))
```

---

# Tables

```{r table-helpers}
# Base CRPS for skill score computation (family-matched)
base_crps_kinsa <- results %>%
  filter(model == "Base + Kinsa") %>%
  select(disease, target_date, horizon, crps_base = crps)

base_crps_ili <- results %>%
  filter(model == "Base + ILI") %>%
  select(disease, target_date, horizon, crps_base = crps)

results_with_skill <- bind_rows(
  results %>%
    filter(str_detect(model, "Kinsa")) %>%
    left_join(base_crps_kinsa, by = c("disease", "target_date", "horizon")),
  results %>%
    filter(str_detect(model, "ILI")) %>%
    left_join(base_crps_ili, by = c("disease", "target_date", "horizon"))
) %>%
  mutate(crpss = 1 - crps / crps_base)

# Table 1 restricted to horizon 0 (nowcast week) — the primary use case
metrics_all <- results_with_skill %>%
  filter(horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(
    n           = n(),
    mean_crps   = mean(crps),
    mean_crpss  = mean(crpss),
    cover_95    = mean(truth >= q025 & truth <= q975) * 100,
    .groups     = "drop"
  )
```

## Table 1: Overall Nowcast Accuracy

```{r table-1}
tbl1 <- metrics_all %>%
  mutate(
    CRPS      = round(mean_crps,  1),
    CRPSS     = round(mean_crpss, 3),
    `95% Cov` = round(cover_95,   0)
  ) %>%
  select(model, disease, CRPS, CRPSS, `95% Cov`)

tbl1_wide <- tbl1 %>%
  pivot_wider(names_from = disease,
              values_from = c(CRPS, CRPSS, `95% Cov`),
              names_glue = "{disease}_{.value}") %>%
  select(model, starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

# Bold best per column
bold_best <- function(x, target = NA) {
  best <- if (is.na(target)) which.min(x) else which.min(abs(x - target))
  out <- as.character(x)
  out[best] <- cell_spec(out[best], bold = TRUE)
  out
}

# Apply bolding column-by-column to avoid earlier columns becoming character
# before later columns are processed
tbl1_fmt <- tbl1_wide
for (col in names(tbl1_fmt)[-1]) {
  if (grepl("Cov", col)) {
    tbl1_fmt[[col]] <- bold_best(tbl1_fmt[[col]], target = 95)
  } else if (grepl("CRPSS", col)) {
    # For CRPSS, higher is better — bold the maximum
    x <- tbl1_fmt[[col]]
    best <- which.max(x)
    out <- as.character(x)
    out[best] <- cell_spec(out[best], bold = TRUE)
    tbl1_fmt[[col]] <- out
  } else {
    tbl1_fmt[[col]] <- bold_best(tbl1_fmt[[col]])  # lower is better
  }
}

kbl(tbl1_fmt, escape = FALSE, align = c("l", rep("r", 9)),
    caption = "Table 1: Nowcast performance at horizon 0 (100-week rolling-origin backtest). CRPS is sample-based from full posterior draws. CRPSS = 1 - CRPS/CRPS\\_base (vs. family baseline; higher is better). Bold = best per column.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  add_header_above(c(" " = 1, "Influenza" = 3, "RSV" = 3, "COVID-19" = 3))
```

## Table 2: CRPSS by Model and Disease

CRPSS = 1 - CRPS_model / CRPS_base. Kinsa models compared against
Base + Kinsa; ILI models compared against Base + ILI. "% Weeks Better"
= percentage of the 100 backtest weeks where this model beat its baseline.

```{r table-2}
skill_tbl <- results_with_skill %>%
  filter(!model %in% c("Base + Kinsa", "Base + ILI"), horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(
    CRPSS      = round(mean(crpss), 3),
    `% Better` = round(mean(crpss > 0) * 100, 0),
    .groups    = "drop"
  )

skill_wide <- skill_tbl %>%
  pivot_wider(names_from = disease,
              values_from = c(CRPSS, `% Better`),
              names_glue = "{disease}_{.value}") %>%
  select(model, starts_with("Influenza"), starts_with("RSV"), starts_with("COVID"))

# Bold highest CRPSS per disease column
skill_fmt <- skill_wide
for (col in names(skill_fmt)[-1]) {
  x <- skill_fmt[[col]]
  if (grepl("CRPSS", col)) {
    best <- which.max(x)
  } else {
    best <- which.max(x)   # % Better: also higher is better
  }
  out <- as.character(x)
  out[best] <- cell_spec(out[best], bold = TRUE)
  skill_fmt[[col]] <- out
}

kbl(skill_fmt, escape = FALSE, align = c("l", rep("r", 6)),
    caption = "Table 2: CRPSS relative to family baseline (Kinsa models vs. Base + Kinsa; ILI models vs. Base + ILI). Higher CRPSS = greater improvement. Bold = best per column. Horizon = 0") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  add_header_above(c(" " = 1, "Influenza" = 2, "RSV" = 2, "COVID-19" = 2))
```

## Table 3: Performance by Horizon

```{r table-3}
horizon_tbl <- results %>%
  group_by(model, disease, horizon) %>%
  summarise(CRPS = round(mean(crps), 1), .groups = "drop") %>%
  mutate(horizon_label = factor(
    horizon, levels = c(-2, -1, 0),
    labels = c("Horizon -2 (2wk back)", "Horizon -1 (1wk back)", "Horizon 0 (nowcast)")
  ))

horizon_wide <- horizon_tbl %>%
  pivot_wider(names_from = disease, values_from = CRPS) %>%
  arrange(horizon, model) %>%
  select(horizon_label, model, Influenza, RSV, `COVID-19`)

kbl(horizon_wide %>% select(-horizon_label),
    align = c("l", rep("r", 3)),
    caption = "Table 3: Mean CRPS by model and horizon. Horizon 0 is the nowcast week (hardest).") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  pack_rows(index = table(horizon_wide$horizon_label))
```

## Table 4: CRPS at Horizon 0 (Nowcast Week)

Sample-based CRPS at horizon 0 only — the hardest and most relevant
prediction task. Kinsa and ILI model families shown side by side.

```{r table-4-crps}
crps_tbl <- results %>%
  filter(horizon == 0) %>%
  group_by(model, disease) %>%
  summarise(CRPS = round(mean(crps), 1), .groups = "drop") %>%
  pivot_wider(names_from = disease, values_from = CRPS) %>%
  mutate(
    family    = if_else(str_detect(model, "Kinsa"), "Kinsa", "ILI"),
    structure = case_when(
      str_detect(model, "Base") ~ "Base",
      str_detect(model, "M1")   ~ "M1",
      str_detect(model, "M2")   ~ "M2",
      str_detect(model, "M3")   ~ "M3"
    )
  ) %>%
  arrange(structure, family) %>%
  select(structure, family, model, Influenza, RSV, `COVID-19`)

# Bold lowest CRPS per (disease, structure) group
crps_fmt <- crps_tbl
for (col in c("Influenza", "RSV", "COVID-19")) {
  vals <- crps_fmt[[col]]
  for (struct in unique(crps_fmt$structure)) {
    idx <- which(crps_fmt$structure == struct)
    best <- idx[which.min(vals[idx])]
    crps_fmt[[col]][best] <- cell_spec(as.character(vals[best]), bold = TRUE)
    crps_fmt[[col]][setdiff(idx, best)] <- as.character(vals[setdiff(idx, best)])
  }
}

kbl(crps_fmt %>% select(-structure),
    escape = FALSE, align = c("l", "l", "r", "r", "r"),
    caption = paste(
      "Table 4: Mean CRPS at horizon 0 (nowcast week) by model and disease.",
      "CRPS is sample-based from full posterior draws (~6,000 samples per fit).",
      "Bold = best within each model structure (Kinsa vs. ILI)."
    )) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE,
                font_size = 12) %>%
  pack_rows(index = table(crps_fmt$structure)) %>%
  add_header_above(c(" " = 2, "Mean CRPS" = 3))
```
