---
title: "Model 2: ILI → Specific Diseases"
output: html_document
date: "2026-03-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rjags)
library(tidyverse)
library(coda)
library(tidycensus)
library(HDInterval)
library(patchwork)
```

```{r}
#census_api_key("81e4817f989c60bf268764565acf7fb7fbe0e2cf", install = TRUE)
```

## Load in the Data
```{r load_data}
data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz")
ili_data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/delphi_ili_fluview/standard/data.csv.gz")


# filtering the data to state and using "total" as the age demographic

fips <- "26" # Michigan

data <- data %>%
  filter(geography == fips , age == "Total")

# Filter ILI data for state
ili_data <- ili_data %>%
  filter(geography == fips) %>%
  select(time, ili_count = delphi_fluview_num_ili, ili_number_patients_tested = delphi_fluview_num_patients) %>%
  arrange(time)

state_pre <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2019,
  time_series = TRUE
) |>
  filter(variable == "POP") |>
  mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |>
  arrange(year)|>
  pull(value)

state_pop <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2025,  # Most recent available
  time_series = TRUE
) |>
  filter(variable == "POPESTIMATE", year %in% c(2020:2025)) %>%
  arrange(year) %>%
  pull(value)


state_pop_estimates <- tibble(
  year = 2018:2026,
  population = c(
    # State census population estimates
    state_pre[1], #2018
    state_pre[2], #2019
    state_pop[1],   #2020
    state_pop[2],   #2021
    state_pop[3],   #2022
    state_pop[4],   #2023
    state_pop[5],   #2024
    state_pop[6],   #2025
    state_pop[6]   #2026 - using 2025 estimate for 2026
  )
)

```

## Merging Population Estimates with Data
```{r population_merge}
data_with_pop <- data %>%
  mutate(year = year(time)) %>%
  left_join(state_pop_estimates, by = "year")

data_with_pop <- data_with_pop %>%
  mutate(
    covid_count = round(rate_covid * population / 100000),
    flu_count = round(rate_flu * population / 100000),
    rsv_count = round(rate_rsv * population / 100000)
  )

# Merge ILI data with the main dataset
data_with_pop <- data_with_pop %>%
  left_join(ili_data, by = "time") %>%
  mutate(ili_count = replace_na(ili_count, 0))
```

## Model 2: ILI → Specific Diseases

ILI is modeled independently, then its log-rate (phi[i,4]) is used as a covariate
in the seasonal means of flu, RSV, and COVID. This leverages ILI's 1-week reporting
advantage: at week T-1, ILI is observed (constraining phi[T-1,4]) while the specific
signals are NA, so the ILI signal directly informs their predictions.

```{r jags-model}
model_string <- "
model {
  # Likelihood - Negative Binomial for all diseases
  for(i in 1:tmax) {
    flu[i]   ~ dnegbin(p1[i], r[1])
    rsv[i]   ~ dnegbin(p2[i], r[2])
    covid[i] ~ dnegbin(p3[i], r[3])
    ili[i]   ~ dnegbin(p4[i], r[4])

    p1[i] <- r[1] / (r[1] + lambda1[i])
    p2[i] <- r[2] / (r[2] + lambda2[i])
    p3[i] <- r[3] / (r[3] + lambda3[i])
    p4[i] <- r[4] / (r[4] + lambda4[i])

    lambda1[i] <- exp(phi[i,1])
    lambda2[i] <- exp(phi[i,2])
    lambda3[i] <- exp(phi[i,3])
    lambda4[i] <- exp(phi[i,4]) * num_patients[i]
  }

  # ILI (k=4) - modeled independently, same as base model
  mu0[4] ~ dnorm(0, 1e-4)
  beta_sin[4] ~ dnorm(0, 1e-4)
  beta_cos[4] ~ dnorm(0, 1e-4)
  prec.phi[4] ~ dgamma(0.01, 0.01)
  rho1[4] ~ dunif(0, 1)

  mu_t[1,4] <- mu0[4] + beta_sin[4] * sin52[1] + beta_cos[4] * cos52[1]
  phi[1,4] ~ dnorm(mu_t[1,4], prec.phi[4] * (1 - pow(rho1[4], 2)))

  for(i in 2:tmax) {
    mu_t[i,4] <- mu0[4] + beta_sin[4] * sin52[i] + beta_cos[4] * cos52[i]
    phi[i,4] ~ dnorm(mu_t[i,4] + rho1[4] * (phi[i-1,4] - mu_t[i-1,4]),
                      prec.phi[4])
  }

  # Flu, RSV, COVID (k = 1, 2, 3) - informed by ILI log-rate
  for(k in 1:3) {
    mu0[k] ~ dnorm(0, 1e-4)
    beta_sin[k] ~ dnorm(0, 1e-4)
    beta_cos[k] ~ dnorm(0, 1e-4)
    prec.phi[k] ~ dgamma(0.01, 0.01)
    rho1[k] ~ dunif(0, 1)
    b_ili[k] ~ dnorm(0, 1e-4)

    mu_t[1,k] <- mu0[k] + beta_sin[k] * sin52[1] + beta_cos[k] * cos52[1]
                 + b_ili[k] * phi[1,4]
    phi[1,k] ~ dnorm(mu_t[1,k], prec.phi[k] * (1 - pow(rho1[k], 2)))

    for(i in 2:tmax) {
      mu_t[i,k] <- mu0[k] + beta_sin[k] * sin52[i] + beta_cos[k] * cos52[i]
                   + b_ili[k] * phi[i,4]
      phi[i,k] ~ dnorm(mu_t[i,k] + rho1[k] * (phi[i-1,k] - mu_t[i-1,k]),
                        prec.phi[k])
    }
  }

  # Overdispersion priors
  r[1] ~ dgamma(0.1, 0.1)
  r[2] ~ dgamma(0.1, 0.1)
  r[3] ~ dgamma(0.1, 0.1)
  r[4] ~ dgamma(0.1, 0.1)
}
"

writeLines(model_string, "model2_ili_to_diseases.jags")
```

## Prepare JAGS Data

```{r jags-data}
# Asymmetric holdout: ILI has 1-week lag, specific signals have 2-week lag
n_holdout_specific <- 2
n_holdout_ili <- 1

tmax <- nrow(data_with_pop)

# Create harmonic vectors for seasonality (52-week cycle)
t_vec <- 1:tmax
sin52 <- sin(2 * pi * t_vec / 52)
cos52 <- cos(2 * pi * t_vec / 52)

cat("Total observations:", tmax, "\n")
cat("Flu/RSV/COVID: training on", tmax - n_holdout_specific, "| evaluating last", n_holdout_specific, "\n")
cat("ILI: training on", tmax - n_holdout_ili, "| evaluating last", n_holdout_ili, "\n")

# Store actual values for comparison BEFORE setting to NA
actuals_holdout_specific <- data_with_pop %>%
  tail(n_holdout_specific) %>%
  select(time, flu_count, rsv_count, covid_count)

actuals_holdout_ili <- data_with_pop %>%
  tail(n_holdout_ili) %>%
  select(time, ili_count)

cat("\nHeld-out actual values (specific signals - last 2 weeks):\n")
print(actuals_holdout_specific)
cat("\nHeld-out actual values (ILI - last 1 week):\n")
print(actuals_holdout_ili)

# Create data vectors with NA for held-out observations
flu_data <- data_with_pop$flu_count
rsv_data <- data_with_pop$rsv_count
covid_data <- data_with_pop$covid_count
ili_data_vec <- data_with_pop$ili_count

# Specific signals: last 2 weeks are NA
flu_data[(tmax - n_holdout_specific + 1):tmax] <- NA
rsv_data[(tmax - n_holdout_specific + 1):tmax] <- NA
covid_data[(tmax - n_holdout_specific + 1):tmax] <- NA

# ILI: only last 1 week is NA (has 1-week advantage)
ili_data_vec[(tmax - n_holdout_ili + 1):tmax] <- NA

# Prepare data list for JAGS
jags_data <- list(
  flu = flu_data,
  rsv = rsv_data,
  covid = covid_data,
  ili = ili_data_vec,
  num_patients = data_with_pop$ili_number_patients_tested,  # ILI exposure offset (never NA)
  tmax = tmax,
  sin52 = sin52,
  cos52 = cos52
)

cat("\nLast few values passed to JAGS (NA = to be predicted):\n")
cat("Flu:", tail(jags_data$flu, 5), "\n")
cat("RSV:", tail(jags_data$rsv, 5), "\n")
cat("COVID:", tail(jags_data$covid, 5), "\n")
cat("ILI:", tail(jags_data$ili, 5), "\n")
```

## Run JAGS Model

```{r run-jags, message=FALSE, warning=FALSE}
# Parameters to monitor - includes b_ili coefficients for ILI-to-disease mixing
params <- c("mu0", "rho1", "prec.phi", "r", "beta_sin", "beta_cos",
            "b_ili",
            "lambda1", "lambda2", "lambda3", "lambda4",
            "flu", "rsv", "covid", "ili")

# Set seeds for reproducibility
inits <- list(
  list(".RNG.seed" = 123, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 456, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 789, ".RNG.name" = "base::Wichmann-Hill")
)

# Initialize the model
jags_model <- jags.model(
  file = "model2_ili_to_diseases.jags",
  data = jags_data,
  n.chains = 3,
  n.adapt = 1000,
  inits = inits
)

# Burn-in
update(jags_model, n.iter = 10000)

samples <- coda.samples(
  model = jags_model,
  variable.names = params,
  n.iter = 20000,
  thin = 10
)
```

## Model Diagnostics

```{r diagnostics}
# Summary of key parameters
summary(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                    "rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]",
                    "r[1]", "r[2]", "r[3]", "r[4]",
                    "beta_sin[1]", "beta_sin[2]", "beta_sin[3]", "beta_sin[4]",
                    "beta_cos[1]", "beta_cos[2]", "beta_cos[3]", "beta_cos[4]",
                    "b_ili[1]", "b_ili[2]", "b_ili[3]")])

# Trace plots for AR(1) parameters
plot(samples[, "rho1[1]"])
plot(samples[, "rho1[2]"])
plot(samples[, "rho1[3]"])
plot(samples[, "rho1[4]"])

plot(samples[, "mu0[1]"])
plot(samples[, "mu0[2]"])
plot(samples[, "mu0[3]"])
plot(samples[, "mu0[4]"])

# Trace plots for overdispersion parameters
plot(samples[, "r[1]"])
plot(samples[, "r[2]"])
plot(samples[, "r[3]"])
plot(samples[, "r[4]"])

# Trace plots for seasonal coefficients
plot(samples[, "beta_sin[1]"])
plot(samples[, "beta_sin[2]"])
plot(samples[, "beta_sin[3]"])
plot(samples[, "beta_sin[4]"])

plot(samples[, "beta_cos[1]"])
plot(samples[, "beta_cos[2]"])
plot(samples[, "beta_cos[3]"])
plot(samples[, "beta_cos[4]"])

# Trace plots for ILI -> disease mixing coefficients
plot(samples[, "b_ili[1]"])  # ILI -> Flu
plot(samples[, "b_ili[2]"])  # ILI -> RSV
plot(samples[, "b_ili[3]"])  # ILI -> COVID

# Gelman-Rubin diagnostic (should be close to 1)
gelman.diag(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                        "rho1[1]", "rho1[2]", "rho1[3]", "rho1[4]",
                        "r[1]", "r[2]", "r[3]", "r[4]",
                        "beta_sin[1]", "beta_sin[2]", "beta_sin[3]", "beta_sin[4]",
                        "beta_cos[1]", "beta_cos[2]", "beta_cos[3]", "beta_cos[4]",
                        "b_ili[1]", "b_ili[2]", "b_ili[3]")])
```
