---
title: "Model 3 (Kinsa): Multivariate Normal AR(1) with Shared Rho"
output: html_document
date: "2026-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rjags)
library(tidyverse)
library(coda)
library(tidycensus)
library(HDInterval)
library(patchwork)
library(scoringRules)
```

```{r}
#census_api_key("81e4817f989c60bf268764565acf7fb7fbe0e2cf", install = TRUE)
```

## Load in the Data

Mirrors `Model_3_MVN_AR1.Rmd` but uses Kinsa `percent_ill` in place of ILI.
Two key differences from the original Model 3:

1. The fourth series (Kinsa) is observed via a lognormal likelihood, not a
   negative binomial.
2. All four latent AR(1) processes share a single autocorrelation
   `rho_shared`, instead of each having its own `rho1[k]`.

```{r load_data}
# -------- FILL IN: local path to your Kinsa CSV --------
kinsa_path <- ""   # e.g. "/Users/tobysalmon/Desktop/SDS-492-Thesis/kinsa.csv"
# -------------------------------------------------------

kinsa_state <- "Michigan"

data <- read_csv("https://raw.githubusercontent.com/PopHIVE/Ingest/main/data/respnet/standard/data.csv.gz")
kinsa_raw <- read_csv(kinsa_path)

fips <- "26" # Michigan

data <- data %>%
  filter(geography == fips , age == "Total")

kinsa_data <- kinsa_raw %>%
  filter(region_name == kinsa_state) %>%
  mutate(time = as.Date(time)) %>%
  select(time, percent_ill) %>%
  arrange(time)

state_pre <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2019,
  time_series = TRUE
) |>
  filter(variable == "POP") |>
  mutate(year = 2009 + DATE) |>
  filter(year %in% c(2018, 2019)) |>
  arrange(year)|>
  pull(value)

state_pop <- get_estimates(
  geography = "state",
  state = fips,
  product = "population",
  vintage = 2025,
  time_series = TRUE
) |>
  filter(variable == "POPESTIMATE", year %in% c(2020:2025)) %>%
  arrange(year) %>%
  pull(value)

state_pop_estimates <- tibble(
  year = 2018:2026,
  population = c(
    state_pre[1], state_pre[2],
    state_pop[1], state_pop[2], state_pop[3],
    state_pop[4], state_pop[5], state_pop[6], state_pop[6]
  )
)
```

## Merging Population Estimates with Data

```{r population_merge}
data_with_pop <- data %>%
  mutate(year = year(time)) %>%
  left_join(state_pop_estimates, by = "year")

data_with_pop <- data_with_pop %>%
  mutate(
    covid_count = round(rate_covid * population / 100000),
    flu_count   = round(rate_flu   * population / 100000),
    rsv_count   = round(rate_rsv   * population / 100000)
  )

data_with_pop <- data_with_pop %>%
  inner_join(kinsa_data, by = "time") %>%
  arrange(time)

stopifnot(all(data_with_pop$percent_ill > 0))
```

## Model 3 (Kinsa): MVN AR(1) with shared rho

All four latent log-scale series evolve jointly via a multivariate normal
AR(1) with a single shared autocorrelation. Innovations across diseases are
correlated through a 4x4 covariance matrix.

```{r jags-model}
model_string <- "
model {
  for(i in 1:tmax) {
    flu[i]   ~ dnegbin(p1[i], r[1])
    rsv[i]   ~ dnegbin(p2[i], r[2])
    covid[i] ~ dnegbin(p3[i], r[3])
    kinsa[i] ~ dlnorm(phi[i,4], tau_kinsa)

    p1[i] <- r[1] / (r[1] + lambda1[i])
    p2[i] <- r[2] / (r[2] + lambda2[i])
    p3[i] <- r[3] / (r[3] + lambda3[i])

    lambda1[i] <- exp(phi[i,1])
    lambda2[i] <- exp(phi[i,2])
    lambda3[i] <- exp(phi[i,3])
    lambda4[i] <- exp(phi[i,4])
  }

  # Seasonal means
  for(k in 1:4) {
    mu0[k] ~ dnorm(0, 1e-4)
    beta_sin[k] ~ dnorm(0, 1e-4)
    beta_cos[k] ~ dnorm(0, 1e-4)

    for(i in 1:tmax) {
      mu_t[i,k] <- mu0[k] + beta_sin[k] * sin52[i] + beta_cos[k] * cos52[i]
    }
  }

  # Single shared AR(1) autocorrelation for all four series
  rho_shared ~ dunif(0, 1)

  # First time point - vague independent priors
  for(k in 1:4) {
    phi[1,k] ~ dnorm(mu_t[1,k], 0.01)
  }

  # Multivariate AR(1) for subsequent time points
  for(i in 2:tmax) {
    for(k in 1:4) {
      mu_phi[i,k] <- mu_t[i,k] + rho_shared * (phi[i-1,k] - mu_t[i-1,k])
    }
    phi[i,1:4] ~ dmnorm(mu_phi[i,1:4], Omega[1:4,1:4])
  }

  # Precision / covariance
  Omega[1:4,1:4] ~ dwish(R[1:4,1:4], 5)
  Sigma[1:4,1:4] <- inverse(Omega[1:4,1:4])

  # Count-process overdispersion
  r[1] ~ dgamma(0.1, 0.1)
  r[2] ~ dgamma(0.1, 0.1)
  r[3] ~ dgamma(0.1, 0.1)

  # Kinsa lognormal precision
  tau_kinsa ~ dgamma(0.01, 0.01)
}
"

writeLines(model_string, "model3_mvn_ar1_kinsa.jags")
```

## Prepare JAGS Data

```{r jags-data}
n_holdout_specific <- 1
n_holdout_kinsa <- 0

tmax <- nrow(data_with_pop)

t_vec <- 1:tmax
sin52 <- sin(2 * pi * t_vec / 52)
cos52 <- cos(2 * pi * t_vec / 52)

cat("Total observations:", tmax, "\n")
cat("Flu/RSV/COVID: training on", tmax - n_holdout_specific, "| evaluating last", n_holdout_specific, "\n")
cat("Kinsa: fully observed\n")

actuals_holdout_specific <- data_with_pop %>%
  tail(n_holdout_specific) %>%
  select(time, flu_count, rsv_count, covid_count)

print(actuals_holdout_specific)

flu_data   <- data_with_pop$flu_count
rsv_data   <- data_with_pop$rsv_count
covid_data <- data_with_pop$covid_count
kinsa_vec  <- data_with_pop$percent_ill

flu_data[(tmax - n_holdout_specific + 1):tmax]   <- NA
rsv_data[(tmax - n_holdout_specific + 1):tmax]   <- NA
covid_data[(tmax - n_holdout_specific + 1):tmax] <- NA

jags_data <- list(
  flu = flu_data,
  rsv = rsv_data,
  covid = covid_data,
  kinsa = kinsa_vec,
  tmax = tmax,
  sin52 = sin52,
  cos52 = cos52,
  R = diag(4)
)
```

## Run JAGS Model

```{r run-jags, message=FALSE, warning=FALSE}
params <- c("mu0", "rho_shared", "r", "tau_kinsa",
            "beta_sin", "beta_cos",
            "Sigma",
            "lambda1", "lambda2", "lambda3", "lambda4",
            "flu", "rsv", "covid", "kinsa")

inits <- list(
  list(".RNG.seed" = 123, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 456, ".RNG.name" = "base::Wichmann-Hill"),
  list(".RNG.seed" = 789, ".RNG.name" = "base::Wichmann-Hill")
)

jags_model <- jags.model(
  file = "model3_mvn_ar1_kinsa.jags",
  data = jags_data,
  n.chains = 3,
  n.adapt = 2000,
  inits = inits
)

update(jags_model, n.iter = 20000)

samples <- coda.samples(
  model = jags_model,
  variable.names = params,
  n.iter = 30000,
  thin = 15
)
```

## Model Diagnostics

```{r diagnostics}
summary(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                    "rho_shared",
                    "r[1]", "r[2]", "r[3]", "tau_kinsa",
                    "beta_sin[1]", "beta_sin[2]", "beta_sin[3]", "beta_sin[4]",
                    "beta_cos[1]", "beta_cos[2]", "beta_cos[3]", "beta_cos[4]")])

cat("\n=== Estimated Covariance Matrix (Sigma) ===\n")
cat("Order: Flu, RSV, COVID, Kinsa\n\n")
summary(samples[, c("Sigma[1,1]", "Sigma[1,2]", "Sigma[1,3]", "Sigma[1,4]",
                    "Sigma[2,2]", "Sigma[2,3]", "Sigma[2,4]",
                    "Sigma[3,3]", "Sigma[3,4]",
                    "Sigma[4,4]")])

plot(samples[, "rho_shared"])
plot(samples[, "mu0[1]"])
plot(samples[, "mu0[2]"])
plot(samples[, "mu0[3]"])
plot(samples[, "mu0[4]"])

plot(samples[, "r[1]"])
plot(samples[, "r[2]"])
plot(samples[, "r[3]"])
plot(samples[, "tau_kinsa"])

# Cross-series covariances with Kinsa (column 4)
plot(samples[, "Sigma[1,4]"])  # Flu-Kinsa
plot(samples[, "Sigma[2,4]"])  # RSV-Kinsa
plot(samples[, "Sigma[3,4]"])  # COVID-Kinsa

gelman.diag(samples[, c("mu0[1]", "mu0[2]", "mu0[3]", "mu0[4]",
                        "rho_shared",
                        "r[1]", "r[2]", "r[3]", "tau_kinsa",
                        "Sigma[1,1]", "Sigma[2,2]", "Sigma[3,3]", "Sigma[4,4]",
                        "Sigma[1,2]", "Sigma[1,3]", "Sigma[1,4]",
                        "Sigma[2,3]", "Sigma[2,4]", "Sigma[3,4]")])
```

## Evaluation

```{r evaluation, warning=FALSE}
samples_combined <- do.call(rbind, samples)

eval_results <- tibble(
  week = 1:n_holdout_specific,
  time = actuals_holdout_specific$time,
  actual_flu   = actuals_holdout_specific$flu_count,
  actual_rsv   = actuals_holdout_specific$rsv_count,
  actual_covid = actuals_holdout_specific$covid_count
)

for (h in 1:n_holdout_specific) {
  idx <- (tmax - n_holdout_specific) + h

  flu_samples   <- samples_combined[, paste0("flu[", idx, "]")]
  rsv_samples   <- samples_combined[, paste0("rsv[", idx, "]")]
  covid_samples <- samples_combined[, paste0("covid[", idx, "]")]

  eval_results$pred_flu_median[h]   <- median(flu_samples)
  eval_results$pred_rsv_median[h]   <- median(rsv_samples)
  eval_results$pred_covid_median[h] <- median(covid_samples)

  eval_results$abs_error_flu[h]   <- abs(eval_results$actual_flu[h]   - median(flu_samples))
  eval_results$abs_error_rsv[h]   <- abs(eval_results$actual_rsv[h]   - median(rsv_samples))
  eval_results$abs_error_covid[h] <- abs(eval_results$actual_covid[h] - median(covid_samples))

  eval_results$crps_flu[h]   <- crps_sample(y = eval_results$actual_flu[h],   dat = flu_samples)
  eval_results$crps_rsv[h]   <- crps_sample(y = eval_results$actual_rsv[h],   dat = rsv_samples)
  eval_results$crps_covid[h] <- crps_sample(y = eval_results$actual_covid[h], dat = covid_samples)
}

print(eval_results)
```

```{r summary-metrics}
metrics <- tibble(
  Disease = c("Flu", "RSV", "COVID"),
  MAE = c(
    mean(eval_results$abs_error_flu),
    mean(eval_results$abs_error_rsv),
    mean(eval_results$abs_error_covid)
  ),
  CRPS = c(
    mean(eval_results$crps_flu),
    mean(eval_results$crps_rsv),
    mean(eval_results$crps_covid)
  )
)

cat("=== Summary Metrics (Model 3 MVN-AR1 + Kinsa, shared rho) ===\n")
print(metrics)
```
